---
title: "Monetary Policy"
subtitle: "Week 12"
author: "Krisna Gupta"
date: "09 May 2021 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: monokai
      highlightLines: true
      countIncrementalSlides: false
    css: ["default","dark.css","dark-fonts.css"]

---

## Last week-ish

- AD-AS see movement of real sector / goods market:
  - Relationship between real GDP and prices
  
- Unemployment gets affected:
  - employment follows GDP: moves in short-run, neutral in long-run.
  - In long-run, wage rate follows prices.
  
- Government can interfere:
  - prices doesn't change in long-run.
  
---

## This week

- We have yet the third market: Money market.

- Money market completes the macroeconomics relationship: growth, unemployment and inflation.

- We will learn supply and demand of money, with interest rate as its price.

- We covered how financial system works.
  - We will then see how central bank fits in.
  
---

class: middle, center

# The money market

<iframe src="https://giphy.com/embed/kvu3mUFYobXqyYEdMY" width="480" height="395" frameBorder="0" class="giphy-embed" allowFullScreen></iframe><p><a href="https://giphy.com/gifs/agenciamango-money-marketing-digital-dinheiro-kvu3mUFYobXqyYEdMY">via GIPHY</a></p>

---

## Demand for money

- We have learned that money is the most liquid form of asset.
  - Easy to use for transaction

- However, there's opportunity cost of holding money
  - money bears little to no return compared to other asset classes.
  
- While there are many returns to asset, we use interest rate as the main opportunity cost of holding money.
  - therefore, interest rate is the 'cost' we have to pay to get the convinience of ease of transaction.

---

## Demand for money

- We can say that interest rate is the price of money.
  
- As interest rate increase, the opportunity cost of holding money increases.

- if interest rate is high, it is better to hold less money and put them in the deposit account.

- remember the lesson last week about money creation:
  - when interest rate is high, borrowing becomes expensive while saving becomes more profitable.
  - This situation leads to hence less money being created.
  
---

## Money demand curve
.pull-left[
```{r educ1,echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
library(dplyr)
library(ggplot2)
library(ggdark)
graph<-ggplot()+
  dark_theme_bw(base_size=22)+
  theme(text = element_text(color='#e0e0e0'),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "#e0e0e0"))
demand <- function(x) 10-x

graph +
  scale_y_continuous(limits=c(0,10), breaks = NULL
                     )+
  scale_x_continuous(limits=c(0,10), breaks = NULL
                     )+
  labs(x="Quantity \n of Money m",y="Interest rate r", title = "")+
  annotate("text", x = 9, y = 2, label = 'MD' ,size=8)+
  stat_function(fun = demand, color="cyan", size=1.1)

```
]

.pull-right[
- money demand is downward sloping

- changes in interest rate leads to movement along the curve:
  - r $\Uparrow$, M $\Downarrow$
  - vice versa
  
- Other changes leads to a shift in the money market
]
---

## What shifts the MD curve?

1. **Changes in aggregate price level**. When price is high, you need more cash in hand to purchase them. When I was young, going out with Rp100k in hand is enough to buy dinner and movie tickets. Now?

1. **Changes in real GDP**. As money is used to buy things, when there are more things to buy, there are more money demanded in the economy.

1. **Changes in institution**. Rules such as reserve requirements matter: higher reserve means less flexibility in creating money.

---

## Money supply

- We have learned that money is created by the banking system.

- However, the main driver of money creation is the central bank.
  - in Indonesia its Bank Indonesia (BI).
  
- Bank Indonesia controls money supply in three ways:
  - Open-market operations via buying and selling treasury bills
  - Discount windows
  - Change reserve requirements.
  
---

## Money supply

Three main tools for BI to control money supply:

1. **Open-market operations** is buying and selling [treasury bill](https://www.bi.go.id/id/fungsi-utama/moneter/operasi-moneter/Default.aspx) (eg Sertifikat Bank Indonesia, SukBI, etc). BI sells t-bill when they want to reduce money supply, and buys them when otherwise.

1. **Discount windows** is opened when banks needed liquidity and BI provide them with interest rate. The interest rate BI charge to banks is called Repo rate ([BI 7DRR](https://www.bi.go.id/id/fungsi-utama/moneter/Default.aspx)). Low repo rate allows banks to expand money supply.

1. **Reserve requirements**. When liquidity is needed, BI will reduce reserve requirements so banks have more money to lend to people. This is the case during [COVID-19 recession](https://www.thejakartapost.com/news/2020/03/02/bi-cuts-reserve-ratio-frees-up-3-2b-liquidity-in-local-banks-amid-market-sell-off.html).

---

## Money market equilibrium
.pull-left[
```{r educ2,echo=FALSE, message=FALSE, warning=FALSE}
rh<-expression(r[H])
re<-expression(r[E])
rl<-expression(r[L])
mh<-expression(m[H])
me<-expression(bar(M))
ml<-expression(m[L])

A<-graph +
  scale_y_continuous(limits=c(0,10), breaks = c(2,5,8),
                     labels=c(rl,re,rh))+
  scale_x_continuous(limits=c(0,10), breaks = c(2,5,8),
                     labels=c(mh,me,ml))+
  labs(x="Quantity \n of Money m",y="Interest rate r", title = "")+
  annotate("text", x = 9, y = 2, label = "MD" ,size=8)+
  annotate("text", x = 6, y = 10, label = "MS" ,size=8)+
  annotate("text", x = 2, y = 9, label = "H" ,size=8)+
  annotate("text", x = 8, y = 3, label = "L" ,size=8)+
  annotate("text", x = 5.5, y = 5.5, label = "E" ,size=8)+
  geom_point(aes(x=5,y=5),color='white',size=5)+
  geom_point(aes(x=2,y=8),color='white',size=5)+
  geom_point(aes(x=8,y=2),color='white',size=5)+
  stat_function(fun = demand, color="cyan", size=1.1)+
  geom_vline(xintercept=5,color="red",size=1.1)+
  geom_segment(aes(x = 0, y = 5, xend = 5, yend = 5), lty = "dotted"
               ,size=2) +
  geom_segment(aes(x = 5, y = 0, xend = 5, yend = 5), lty = "dotted"
               ,size=2)+
  geom_segment(aes(x = 0, y = 8, xend = 2, yend = 8), lty = "dotted"
               ,size=2) +
  geom_segment(aes(x = 2, y = 0, xend = 2, yend = 8), lty = "dotted"
               ,size=2)+
  geom_segment(aes(x = 0, y = 2, xend = 8, yend = 2), lty = "dotted"
               ,size=2) +
  geom_segment(aes(x = 8, y = 0, xend = 8, yend = 2), lty = "dotted"
               ,size=2)
A
```
]

.pull-right[
- This is called **liquidity preference model of interest rate**.

- BI sets how much money in the market by setting interest rate. Effectively, BI controls the movement of MS.

- When BI wants less money circulated, they crank-up interest rate by using its three tools.


]
---

## Money market equilibrium
.pull-left[
```{r educ3,echo=FALSE, message=FALSE, warning=FALSE}
A
```
]

.pull-right[
- Suppose economy booms and demand for money at $L$, but BI keeps MS at $\overline{M}$.

- At $L$, money demand is $M_L$ and interest rate is at $r_L$

- For those who want to sell non-money asset, they has to offer higher interest rate.
- Which cranks up general interest rate and reduce money demand back to $\overline{M}$
]

---

## Changing money supply
.pull-left[
```{r educ4,echo=FALSE, message=FALSE, warning=FALSE}
m1<-expression(bar(M)[1])
m2<-expression(bar(M)[2])
r1<-expression(r[1])
r2<-expression(r[2])


B<-graph +
  scale_y_continuous(limits=c(0,10), breaks = c(3,7),
                     labels=c(r2,r1))+
  scale_x_continuous(limits=c(0,10), breaks = c(3,7),
                     labels=c(m1,m2))+
  labs(x="Quantity \n of Money m",y="Interest rate r", title = "")+
  annotate("text", x = 9, y = 2, label = "MD" ,size=8)+
  annotate("text", x = 4, y = 10, label = expression(MS[1]) ,size=8)+
  annotate("text", x = 8, y = 10, label = expression(MS[2]) ,size=8)+
  annotate("text", x = 3.5, y = 7.5, label = expression(E[1]) ,size=8)+
  annotate("text", x = 7.5, y = 3.5, label = expression(E[2]) ,size=8)+
  geom_point(aes(x=3,y=7),color='white',size=5)+
  geom_point(aes(x=7,y=3),color='white',size=5)+
  stat_function(fun = demand, color="cyan", size=1.1)+
  geom_vline(xintercept=3,color="red",size=1.1,linetype="dashed")+
  geom_vline(xintercept=7,color="red",size=1.1)+
  geom_segment(aes(x = 0, y = 7, xend = 3, yend = 7), lty = "dotted"
               ,size=2) +
  geom_segment(aes(x = 0, y = 3, xend = 7, yend = 3), lty = "dotted"
               ,size=2)+
  annotate("segment", x = 3.5, xend = 6.5, y = 9, yend = 9, 
           colour = '#E6E6E6', size=2, alpha=0.6, arrow=arrow())+
  annotate("segment", x = 3.7, xend = 6.5, y = 6.7, yend = 4, 
           colour = '#E6E6E6', size=2, alpha=0.6, arrow=arrow())
B
```
]

.pull-right[
- Suppose the cenral bank issue **expansionary monetary policy** either by purchasing t-bill or reducing repo rate.

- The policy pushes money to the banking system, which leads banks to reduce their interest rate.

- Lower interest rate lower borrowing costs, create more money in circulation.
]
---

class: middle, center

# Monetary policy and Aggregate Demand

---

## Monetary policy and Aggregate Demand

- In the previous week, we touched a bit on what moves AD curves.

- When BI increase money supply:
  - interest rate drops
  - low interest rate = low borrowing costs:
    - investment and consumption increases.
    - leads to AD shifts to the right.
    
--

- When BI decrease money supply:
  - interest rate increases
  - high interest rate = high borrowing costs:
    - people saves; industries halt investment.
    - leads to AD shifts to the left.

---

## Monetary policy

- Central bank should only reduce interest rate when output is below potential output.

- The reason is the same as the government:
  - during normal times, boosting AD will results in inflation.
  
- In fact, inflation is the main mandate of BI, NOT output nor unemployment.

--

- The main mandate of BI is to maintain the value of IDR:
  - controlled inflation
  - controlled exchange rate
  
---

## Inflation targeting Framework

- Inflation targeting framework (ITF) is a framework where the basis of BI's monetary policy is inflation.

- BI announce its inflation target (together with Ministry of Finance), and vows to use monetary policy to get that inflation rate.
  - as of [today](https://www.bi.go.id/id/statistik/indikator/target-inflasi.aspx), the target is 3$\pm$1%
  
- When inflation is below target, BI will be expansive, whereas high inflation will leads BI to rise interest rate.

- Indonesia is an open economy: inflation can happen due to cheap IDR. This is why BI also intervene in the exchange rate market.

---

```{r inflasi,echo=FALSE, message=FALSE, warning=FALSE}

inflasi <- data.frame(
  "Year"=seq(2006,2019,1),
  "inflation"=c(6.6,6.59,11.06,2.78,6.96,3.79,4.3,8.38,8.36,3.35,3.02,3.61,3.13,2.72),
  "inflationn"=c(8,6,5,4.5,5,5,4.5,4.5,4.5,4,4,4,3.5,3.5),
  "birate"=c(9.75,8,9.25,6.5,6.5,6,5.75,7.5,7.75,7.5,4.75,4.25,6,5)
)

inflasi %>%
  kbl(align = "r", format='html',
    col.names = c("Year","Actual \n Inflation", "Target \n Inflation","End of Year \n BI Rate"),digits = 2) %>%
  kable_styling(font_size = 18,full_width = F) %>%
  column_spec(1, width = "5cm") %>%
  add_footnote("Source: BI")
```

When inflation was high, BI rate is also high. High BI rate is supposed to have money absorption effect.

---

## Expansionary monetary policy

.pull-left[
```{r lr,echo=FALSE, message=FALSE, warning=FALSE}
demand1 <- function(x) 12-x
supply1<-function(x) x-2
supply <- function(x) x
demand <- function(x) 10-x
graph +
  scale_y_continuous(limits=c(0,10), breaks = c(5,6),
                     labels=c(expression(P[1]),expression(P[2])))+
  scale_x_continuous(limits=c(1,11), breaks = c(5,6),
                     labels=c(expression(Y[1]),expression(Y[2])))+
  labs(x="Real GDP",y="Aggregate price \n (GDP deflator)", title = "Positive demand shock in SR")+
  annotate("text", x = 9, y = 8, label = 'SRAS' ,size=8)+
  annotate("text", x = 9, y = 2, label = expression(AD[1]) ,size=8)+
  annotate("text", x = 10, y = 3, label = expression(AD[2]) ,size=8)+
  annotate("text", x = 6, y = 5, label =expression(E[1]) ,size=8)+
  geom_point(aes(x=5,y=5),color='white',size=5)+
  annotate("text", x = 7, y = 6, label =expression(E[2]) ,size=8)+
  geom_point(aes(x=6,y=6),color='white',size=5)+
  stat_function(fun = supply, color="red", size=1.1)+
  stat_function(fun = demand, color="cyan", size=1.1, lty=3)+
  stat_function(fun = demand1, color="cyan", size=1.1)+
  geom_segment(aes(x = 1, y = 5, xend = 5, yend = 5), lty = "dotted"
               ,size=2) +
  geom_segment(aes(x = 5, y = 0, xend = 5, yend = 5), lty = "dotted"
               ,size=2)+
  geom_segment(aes(x = 1, y = 6, xend = 6, yend = 6), lty = "dotted"
               ,size=2) +
  geom_segment(aes(x = 6, y = 0, xend = 6, yend = 6), lty = "dotted"
               ,size=2)
```
]

.pull-right[
- When the central bank increase money supply, interest rate goes down, leads to inrease investment and consumption.

- AD shifts to the right, output go up to $Y_2$ while aggregate price level increases to $P_2$

]
---

## Inflation shifts MD...

.pull-left[
```{r grafik1,echo=FALSE, message=FALSE, warning=FALSE}
demand2 <- function(x) 14-x
B +
  stat_function(fun = demand2, color="cyan", size=1.1)+
  geom_segment(aes(x = 0, y = 7, xend = 7, yend = 7), lty = "dotted"
               ,size=2)+
  annotate("text", x = 7.3, y = 7.3, label =expression(E[3]) ,size=8)+
  annotate("segment", x = 7.4, xend = 7.4, y = 3, yend = 7, 
           colour = '#E6E6E6', size=2, alpha=0.6, arrow=arrow())+
  annotate("text", x = 9, y = 5, label = expression(MD[2]) ,size=8)+
  geom_point(aes(x=7,y=7),color='white',size=5)
```
]

.pull-right[
- The central bank can either pull money supply back, or keep money supply high.

- If money supply stay high, because aggregate demand and aggregate price both go up, MD shifts to the right, increasing money demand.

- In the long-run, this push interest rate back up to the original position.
]
---

## Long run impact

.pull-left[
```{r lr2,echo=FALSE, message=FALSE, warning=FALSE}
demand1 <- function(x) 12-x
supply1<-function(x) x+2
supply <- function(x) x
demand <- function(x) 10-x
graph +
  scale_y_continuous(limits=c(0,10), breaks = c(5,6,7),                                      labels=c(expression(P[1]),expression(P[2]),
                                                                                                      expression(P[3])))+
  scale_x_continuous(limits=c(1,11), breaks = c(5,6),
                     labels=c(expression(Y[1]),expression(Y[2])))+
  labs(x="Real GDP",y="Aggregate price \n (GDP deflator)", title = "Positive demand shock in SR")+
  annotate("text", x = 9, y = 8, label = expression(SRAS[1]) ,size=8)+
  annotate("text", x = 7.5, y = 9, label = expression(SRAS[2]) ,size=8)+
  annotate("text", x = 9, y = 2, label = expression(AD[1]) ,size=8)+
  annotate("text", x = 10, y = 3, label = expression(AD[2]) ,size=8)+
  annotate("text", x = 6, y = 5, label =expression(E[1]) ,size=8)+
  geom_point(aes(x=5,y=5),color='white',size=5)+
  annotate("text", x = 7, y = 6, label =expression(E[2]) ,size=8)+
  geom_point(aes(x=6,y=6),color='white',size=5)+
  annotate("text", x = 6, y = 8, label =expression(E[3]) ,size=8)+
  geom_point(aes(x=5,y=7),color='white',size=5)+
  stat_function(fun = supply, color="red", size=1.1,lty=3)+
  stat_function(fun = supply1, color="red", size=1.1)+
  stat_function(fun = demand, color="cyan", size=1.1, lty=3)+
  stat_function(fun = demand1, color="cyan", size=1.1)+
  geom_vline(xintercept=5,color="orange",size=1.1)+
  geom_segment(aes(x = 1, y = 5, xend = 5, yend = 5), lty = "dotted"
               ,size=2) +
  geom_segment(aes(x = 5, y = 0, xend = 5, yend = 5), lty = "dotted"
               ,size=2)+
  geom_segment(aes(x = 1, y = 6, xend = 6, yend = 6), lty = "dotted"
               ,size=2) +
  geom_segment(aes(x = 6, y = 0, xend = 6, yend = 6), lty = "dotted"
               ,size=2)
```
]

.pull-right[
- Increase of interest rate and wage rate in the long run leads to lower supply (SRAS shifts to the left)

- In the end, output level is back to normal, but prices go up (inflation)

- The central bank failed to keep GDP and employment high, sacrifices its main target which is inflation.

]

---

## Long-run neutrality of money

- Often, many government asks the central bank to finance their expansive program instead of stabilizing prices.

- But when the economy is in its normal state, expansionary monetary policy will only leads to higher inflation.

- This is called **neutrality of money**, that is, during normal times, you can't boost real GDP because increased prices will neutralize the higher amount of money.

---

## Importance of ITF

- The central bank (and the government) have very important job in mitigating recession and stabilizing the economy.

- It is not easy, however:
  - Tightening monetary and fiscal policy too soon risks economy return to recession.
  - Tightening monetary and fiscal policy too late, and the economy may overheat, inflate, and bursting bubbles.
  
- That is why inflation targeting is important, as inflation is the sign of overheating economy.

---

## In short...

**During a recession:**

1. Self-correcting mechanism:
  - Short-run: AD shifts to left, prices go down, output/GDP go down.
  - Long-run, self-correcting mechanism: AS shifts to right, prices go even lower but output returns to normal.
  - In the end, output stays the same, prices level is very low.

1. Government intervention:
  - Short-run: AD shifts to left, prices go down, output/GDP go down.
  - Government interventions: AD shifts to the right, prices and output go back to normal.
  - In the end, output and prices is the same.

---

## Does monetary policy work?

- Note that the central bank does not dictate interest rate charged by private banks.

- BI set repo rate to increase incentive for banks to expand their lending.

- However, there are possibilities that consumer banks do not follow the rate set by BI rate.

--

- One reason is **high risk premia**. Some countries has terrible investing condition. In this situation, default rate may be too big for banks to lower their interest rate and forego their reserves.

--

- Second reason is **oligopolistic banking market**. Remember that interest rate is basically the price of lending service. Banks can hold high interest rate to maximize profit the same way normal monopoly and oligopoly works.

---

## Does monetary policy work?

- Banking sector can become oligopolistic when there are high barriers to entry.

- When barriers to entry to banking is high but demand for lending exist, illegal lending pops-up, e.g.:
  - *lintah darat*
  - illegal Fin-tech

- Third problem is what plague advance countries: **zero-lower bound**. In advance countries, interest rate is so low, it has to go negative to be lowered.

- De facto wise, interest rate is already negative in some countries if you consider admin fee. But we still use banks amid convenience.

---

## What to do?

- When monetary policy isn't effective, the central bank can turn to the government.

- Instead of selling t-bill to banks, BI can sell t-bill straight to the government to finance $G$

- This effectively turns monetary policy to fiscal policy.

- However, this method risks hyper-inflation as in 1966. In fact, since it is in the government's interest to grow GDP, it can ask central bank to purchase t-bill even during normal times.

---

## Next-up

- Macroeconomics essentially discussing relationship between aggregate demand, aggregate supply, aggregate prices, unemployment, interest rate and money supply.

- These understandings are important to make sense of macroeconomic news.

- However, we have not discuss one crucial thing: international economics.

- Next-week we will learn one more important indicator: balance of trade and balance of payment.

- Further, we will see how exchange rate and international credit market plays a role in the economy.

---

class: middle, center

# The end