---
title: "Oligopoly & Externality"
subtitle: "Week 7"
author: "Krisna Gupta"
date: "22 March 2021 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: monokai
      highlightLines: true
      countIncrementalSlides: false
    css: ["default","dark.css","dark-fonts.css"]

---

## Announcement

- No quiz again for today

- your assignments = 4 quiz

- Take your time and no need to do all 4 if you are confident.

- Your mid-term will cover week 1 to week 7 (today)

---

## Score so far

###### note: This slide will be deleted after today's class

| no | Q1 | Q2 | Q3 | Q4 | Q5 |
| --- | --- | --- | --- | --- | --- |
| c13200031 | 100 |	80 | 80 | 80 | 50 |
| c13200042 | 100	|80|	50|	70|	60 |
| c13200063 | 100	|90	|90	|80	|90|
| c13200066 | 100	|80	|70	|50	|50 |

.s[
- Remember, I only take best of 10.
    + If you do all 4 Qs on the assignment, you will have 9 scores.
    + More quizzes and assignments after mid-term exam.
]

---

## Recap on last week

- We have learned the two extremens: perfect competition & Monopoly

- In perfect competition, profit only last in the short run. As more firms join the market, price pushed down.

- As long as there are profit, more firms will enter the market until  zero profit is reached.

- In the monopoly, profit maximization can last forever because no more firm can join in.

---

## Today's content

- The imperfect competition: oligopoly and monopolistic competition.
  
- A touch of externality ends the mid-term content.

---

class: center, middle

# Oligopoly

---

## Skutik Price Cartel

- In 2017, AHM and YIMM, two of the biggest skutik makers, were charged by KPPU for an alleged [price fixing](https://www.thejakartapost.com/news/2019/05/02/supreme-court-rejects-appeal-to-verdict-on-yamaha-honda-price-cartel.html).

- In the perfect market, many producers are competing with each other, while in the monopoly, well, there's no other producer.

- If the market is shared by small number of producers, they can decide a price together to behave similar to a monopoly.
  - this is called price fixing.
  
---

### A hypothetical skutik market

.pull-left[
```{r tab1,echo=FALSE, message=FALSE, warning=FALSE}
library(kableExtra)
library(dplyr)
library(ggplot2)
library(ggdark)

graph<-ggplot()+
  dark_theme_bw(base_size=25)+
  theme(text = element_text(color='#e0e0e0'),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "#e0e0e0"))

P<-seq(17,7,length=11)
Q<-seq(100,600,50)
C<-rep(7,11)
profit<-(P-C)*Q

skutik<-data.frame('P'=P,'Q'=Q,'Profit'=profit)

skutab<-kbl(skutik, format='html', align='c') %>%
  add_header_above(c('million IDR','1000 unit','million IDR')) %>%
  add_header_above(c('Unit Price','Monthly sales','Total Revenue')) %>%
  kable_styling(font_size = 16)
skutab
```
]

.pull-right[.s[
- Suppose skutik market behave like this table.
- As you notice, increasing price is associated with decreasing demand.

- Any monopolist operate in the skutik market have that revenue.

- To simplify things, suppose per-unit cost is stagnant at 7 million IDR.

- A monopolist will set P=13, produce Q=300 and gets 1800 as profit

]]

---

### A 2-firm competition

.pull-left[
```{r tab2,echo=FALSE, message=FALSE, warning=FALSE}
skutab
```
]

.pull-right[.s[
- Suppose another firm enters the market and offering P=12.

- Everyone will buy from the entrant.

- the incumbent will have zero sales while the new entrant will get 1750 profit.

- Since per unit cost is 7, the incumbent will offer a discount to match the price.

- Eventually, P=7
]]

---

### A duopoly

.pull-left[
```{r tab3,echo=FALSE, message=FALSE, warning=FALSE}
skutab
```
]

.pull-right[.s[
- A strategic firms know this. They will fare better if they collude and form a **cartel**.

- Both can settle a **price-fixing** deal to limit their Q=150 and set the market price to P=13.

- As you can see, this is a monopoly price and quantity.

- Each will get profit=900, which is much higher than if they compete.
]]

---

## A Duopoly

- A duopoly has a potential to produce a similar outcome as a monopoly.

- As the example above, the duopoly market produces only 300 with price as high as 13, while a competition makes 600 where P=7.

- In short, like a monopoly, duopoly also harm consumers and is inefficient.

- However, a duopoly is harder to maintain because every oligopolist has an incentive to deviate.

---

## Deviation

.pull-left[
```{r tab4,echo=FALSE, message=FALSE, warning=FALSE}
skutab
```
]

.pull-right[.s[
- Suppose both producers agreed to produce Q=150 each at P=13

- However, the incumbent decide that it will produce Q=200 instead.

- Total Q=350 and P=12

- Incumbent's profit= $200 \times 5 = 1000$

- Entrant's profit= $150 \times 5 = 750$
]]
---

## Non-cooperative behavior

- Any player in a duopoly have an incentive to behave **non-cooperatively**.

- By being non-cooperative, the incumbent improve its profit from 900 to 1000 at the expense of the entrant's profit.

- If the entrant saw this coming, it will behave the same way. It is easy to see how a **cartel** can break down on its own.

- This is the reason why an industry with an oligopolistic structure can have a less problematic market outcome than a monopoly in the real world.

---

## Game played by oligopolists

- Oligopolist realize their profits depend on other players and vice-versa.

- This situation is called **profit interdependence** where your own profit depends on the action of others in the market.
  + No interdependence in perfect competition.
  + monopoly is a single player

- essentially oligopolists play a "game" in which the success of a player depends on other players' action.

- This branch of economics is called **game theory**.

---

## Game theory

- Let us represent the previous case with a game theory framework.

- There are 2 **players**: AHM and YIMM.

- There are 2 **actions** each player can take: Produces 150.000 units of skutik or 200.000

- There are 4 possible sets of **payoff**, or profit scenario.

- This game can be represented with a 2 $\times$ 2 matrix. A table of game theory matrix is called **payoff matrix**.

---

### Payoff matrix

```{r game,echo=FALSE, message=FALSE, warning=FALSE}
game <- data.frame(
  "P1"=c(rep('AHM',times=2)), "P2"=c("Q=150","Q=200"),
  "HP"=c("(900,900)","(1000,750)"), "LP"=c("(750,1000)","(875,875)")
)

matrix<-kbl(game, align = "c",format='html',
    col.names = c("","","Q=150","Q=200")) %>%
  collapse_rows(columns = 1) %>%
  add_header_above(c(" "," ", "YIMM" = 2)) %>%
  kable_material()
matrix
```

- each element of a payoff matrix consists of a profit from two players, hence two numbers.

- Player 1 (AHM, left) vs player 2 (top, right)

- payoff structure is (player 1,player 2) for every element.

---

### Payoff matrix

```{r game1,echo=FALSE, message=FALSE, warning=FALSE}
matrix
```

- For example, the top left payoff element means both players get 900.

- Top right payoff element means AHM gets 750 while YIMM gets 1000.

- etc

---

### Payoff matrix

```{r game2,echo=FALSE, message=FALSE, warning=FALSE}
matrix
```

- We get the situation on the top left when both players play (Q=150,Q=150).

- We get the situation on the top right when players play (Q=150,Q=200)

- In Which  AHM set Q=150, YIMM set Q=200

---

### Dominant strategy

```{r game3,echo=FALSE, message=FALSE, warning=FALSE}
matrix
```

- You analize each player separately to see which strategy they will pick by comparing their payoff in different situation.

- If you are AHM, which action you pick when YIMM pick Q=150? Which action you pick when YIMM pick Q=150?

---

### Dominant strategy

```{r game4,echo=FALSE, message=FALSE, warning=FALSE}
matrix
```

- If you are AHM, Q=200 is called **dominant strategy** because no matter what YIMM does, Q=200 give the best outcome.

- Similarly, Q=200 is a dominant strategy for YIMM.

---

### Nash Equilibrium

```{r game5,echo=FALSE, message=FALSE, warning=FALSE}
game2 <- data.frame(
  "P1"=c(rep('AHM',times=2)), "P2"=c("Q=150","Q=200"),
  "HP"=c("(900,900)","(1000,750)"), "LP"=c("(750,1000)","(875,875)*")
)

matrix2<-kbl(game2, align = "c",format='html',
    col.names = c("","","Q=150","Q=200")) %>%
  collapse_rows(columns = 1) %>%
  add_header_above(c(" "," ", "YIMM" = 2)) %>%
  kable_material()
matrix2
  
```

- Since both have Q=200 as a dominant strategy, then we called (Q=200,Q=200) as a **Nash Equilibrium**.

- A **set of actions** is called Nash Equilibrium (NE) when nobody gains from deviating from that set.

---

### Nash Equilibrium

```{r game6,echo=FALSE, message=FALSE, warning=FALSE}
game2 <- data.frame(
  "P1"=c(rep('AHM',times=2)), "P2"=c("Q=150","Q=200"),
  "HP"=c("(900,900)","(1000,750)"), "LP"=c("(750,1000)","(875,875)*")
)

matrix2<-kbl(game2, align = "c",format='html',
    col.names = c("","","Q=150","Q=200")) %>%
  collapse_rows(columns = 1) %>%
  add_header_above(c(" "," ", "YIMM" = 2)) %>%
  kable_material()
matrix2
  
```

- To analyze an NE, you can start from each possible outcome and see whether deviating gives a better payoff to anyone.

- (Q=200,Q=200) is an NE because deviating gives a worse payoff for both

---

### Prisoner's Dilemma

- This payoff structure is a typical of **prisoner's dilemma**

- A game is called prisoner's dilemma when each player has an incentive to **cheat** regardless of what others do, but when both cheat, both are worse off.

- That is why it is important for every players in this type a game to communicate with each other, but it is against the law in many countries.

- Making a written contract is even worse.

---

### Tacit collusion

- It is possible to signal another player without having to meet each other.

- AHM can try to set price high and see if YIMM would follow with a high price, signalling collusion.

- If YIMM followed with a low price, it is easy for AHM to punish YIMM's behavior by making a discount, hence both ended up at the NE.

- Setting price is not against the law.

- this behavior is called **tacit collusion**.

---

### Dealing with oligopolies

- It is even possible for a firm to mergers or making an acquisition to strengthen their market power.

- That's why governments often scrutinize mergers and acquisition. Sometimes even force big firms to separate their business.

- Perhaps the best way to deal with oligopolies is to introduce competition.
  + Lion group and GIA once considered an oligopolies until Air Asia come in.
  + foreign comptition also improved telco market.
  
---

### Dealing with oligopolies

- Sometimes it is hard because an industry is a monopoly by nature due to economies of scale and/or network effect.

- Wholesales, cinema, social media, telco, you name it.

- small producers can be considered oligopolies if operate in a small market.
  - Two minimarkets operate in a small town can be considered a duopoly.
  - definition of market matters, again.
  
---

class: center, middle

# Monopolistic competition

---

### Monopolistic competition

- Monopolistic competition is also a type of an imperfect market, in which a seller may have a control over price.

- It still have characteristics of a perfectly competitive industry, with only differnce is the goods is differentiated.

- The differentiation is not so much that it still have a close substitute.

---

### Example of a product differentiation

- In a food court consists of 10 stalls, you might see different food types being offered.
  + maybe 1 stall is Bakmi GM, another is KFC, Chinese food, italian food, Pecel, drinks, etc.

- Each stall have a bit of control over their price because they have a specialized food.

- However, charge too much and people will flee because other stalls are still food.

  + if a portion of KFC is 100.000 IDR, I won't mind eating bakmi GM everday.

---

### Monopolistic competition

- Suppose all 10 stalls offer Nasi Goreng with exact same quality, then it become a perfect competition where sellers will compete only on price.

- Competing by making a different product can be called **blue ocean strategy**, while competing over cost can be called **red ocean strategy**.

- blue ocean because you create your own market base, red ocean because you compete in an already crowded market and fight over a limited share of market.

---

### Product differentiation

Monopolistic competition is driven by product differentiation:

1. differentiation by **style/type** like our food stall example. Other example: specializing in women's clothing, or in utility vehicle.

2. differentiation by **location**: all indomaret sell the same goods, but we prefer one that's near our location.

3. differentiation by **quality**: there's $10 shoes, and there's $100 shoes.

---

### Product differentiation

- Product differentiation can be annoying: streaming services, apple vs android vs windows

- However, we generally don't hate monopolistic competition because:

  + it improves variety in the market. (imagine if all cars are Avanza)
  
  + Cater everyone's need (not all people buys cheap or expensive products)

  + Competition still push producers to innovate.

---

class: center, middle

#externalities

---

### Externalities

- Is another problem for perfect market because it creates **interdependence**, that is buyer's or seller's individual action affects the whole market.

- The best example is **pollution**. A firm's decision to lower its **own cost** by using high emission power plant imposes cost to **everyone else**.

- Traffic is another good example: a car user contributes to a traffic jam and has no incentive to care about other road users.

- Smoking causes bad health to smokers but also impose cost to everyone around the smoker. etc

---

### Externalities

- All of the examples on the previous slides are **negative externalities**.

- An example for **positive externalities**:

  + Having a bee farm and an apple tree farm leads to higher production because bees help pollinates apple tree.
  + Having smart kids help not only a family but also the whole country.
  
- Externalities also sometimes called **spillover effects**

---

### Externalities 
  
- The problem with externalities is that the individual who do the action does not take into account the social cost and social benefit of their action.

- That's why government would like to limit the activities that cause social cost.
  + this is typicall done via taxation.

- On the other hand, activities that cause positive spillovers needs to be promoted / subsidized.

---

<iframe src="https://embed.polleverywhere.com/free_text_polls/3lq05WgaODztTrjrByiQj?controls=none&short_poll=true" width="800px" height="600px"></iframe>

---

<iframe src="https://embed.polleverywhere.com/free_text_polls/AdfFsnUyUcaQnPOQ9lblT?controls=none&short_poll=true" width="800px" height="600px"></iframe>

---

### Externalities & Common goods

- externalities can be mixed up with a **common good**.

- A common is a type of good that can be consumed by everyone, but a consumption from a person leads to less good for others to consume.

- This also creates **interdependence**: someone's consumption affects consumption of others in the economy.

- An example is a ground water, a meadow, or fish stocks.

---

### Tragedy of the commons.

- In a housing complex, people will make a ground water well for themselves.

- However, ground water has limitation on how fast it recovers.

- If more people in a housing complex make ground water well, there will be less water for others.

- There is no incentive to preserve ground water individually, especially if you know your neighbours will also build ground water.

- In the end, everyone consume unsustainably.

---

### Tragedy of the commons.

- This is called **tragedy of the commons**, a situation where everyone consume even more for their own benefit.

- Another example is fish stock. If you know other fisherman will fish unsustainably, you might fish even more unsustainably to reap the benefit to yourself.

- Like negative externalities, government usually tries to intervene to preserve common goods since it won't solve individually.

- Tax or quota are often use to deal with this situation.

---

### Interdependence

- In a perfect market, acting based on self-interest gives best outcome for the economy.

- However, interdependence creates problem because acting based on self-interest may harm the economy overall.

- Imperfect market: oligopoly.

- Interdependence causes market failure: externalities & tragedy of the commons.

- Unfotunately dealing with these problems are not easy. It is important for you to understand how problems arise from interdependence.

---

# The end of part 1

What we've learned so far:

- What is economics.

- Perfect competition: how supply demand shape the market.

- Market distortion: tax, price and quota policy.

- The importance of elasticities.

- Costs and how firms shape the market.

- Market structure & interdependence

---

# The end of part 1

- This is the end the first part of our class.

- Mid-term exam will only cover up to today's content.

- Starting next week is part 2 which is final exam materials.