---
title: "Ricardian Model"
subtitle: "ECES905205 pertemuan 2"
date: "5 September 2022"
author: "I Made Krisna Gupta"
format: 
  revealjs:
    theme: [simple,../mytheme.scss]
    chalkboard: true
    incremental: false
    transition: convex
    background-transition: fade
    slide-number: true
  beamer: default
---

## Reminder

- Your lecturer pre-midterm is I Made Krisna aka [Imed](https://krisna.or.id).

    - contents are mostly to catch you up with trade theories.

- Kiki Verico will handle post-midterm.

- You'll have biweekly assignments. Please do it as best as you can.

- Everything is on EMAS (at least pre-midterm).

## Learning outcomes

- Last week you've learned the new challenges in globalization & Indonesia's trade pattern.

Today we learn Ricardian model of trade (aka comparative advantage) so:

- you can explain how the model works

- show where gains from trade come from

## Today

- Ricardian setting

- Trade and its gais.

## Concepts

Concepts in trade is straightforward but people always lost idk why

- Differences across countries are a key reason why trade occurs.

- Economies of scale can also be a strong reason.

- Opportunity cost is the root of comparative advantage concept which is the core of all trade model.

## Comparative Advantage

- The opportunity cost of producing something measures the cost of not being able to produce something else with the resources used.

- Comparative advantage will be determined by comparing opportunity costs across countries.

> A country $j$ has a comparative advantage in producing good $i$ if the opportunity cost of producing good $i$ is lower in the country $j$ than in other countries.

## One factor economy {.m}

1. Labor is the only factor of production.
1. Labor productivity varies across countries due to differences in technology, but labor productivity in each country is constant.
1. The supply of labor in each country is fixed.
1. Two goods: food and cloth.
1. Competition allows workers to be paid a wage equal to the value of what they produce, and allows them to work in the industry that pays the higher wage.
1. Two countries: home and foreign.

## One factor economy

- Labor supply $L$ is the total labor resources a country has.

- $a_{LF}$ indicates the amount of labor required to produced 1 kg of Food.

- Food production $Q_F$=total food produced by the home country.

- Production possibility frontier (PPF): the maximum amount of production given resources.

$$
a_{LF}Q_F+a_{LC}Q_C\leq L
$$

## PPF

$$
a_{LF}Q_F+a_{LC}Q_C\leq L
$$

- If $Q_C$=0, maximum food production is $Q_F=\frac{L}{a_{LF}}$

- If $Q_F$=0, maximum cloth production is $Q_C=\frac{L}{a_{LC}}$

## PPF ex.

- Say home country have this spec:

$$
Q_F+2Q_C \leq 1000
$$

- How many labor needed to make 1 cloth? How many 1 food?

- Max possible food is 1000 kg, max possible cloth is 500 metre.

## Home PPF

::: {.s layout-ncol="2"}

- The production function has a linear PPF.

- Home can produce anywhere on the line.

    - below the line $\rightarrow$ inefficient
    
    - above the line $\rightarrow$ unfeasible

- As you move along the line, you increase a production of 1 good while decreasing another.

    - the opportunity cost of making a good is the other good!

```{r}
#| fig-width: 5
#| fig-height: 4
#| label: fig1
#| fig-cap: "Production Possibility Frontier (PPF) country H"
library('ggplot2')
graph<-ggplot()+
  theme(panel.background = element_rect(fill = "#f0f1eb",
                                        colour = "#f0f1eb"),
        plot.background = element_rect(fill = "#f0f1eb"),
        legend.key = element_rect(fill = "#f0f1eb",linetype = "blank"),
        legend.background = element_rect(fill="#f0f1eb"))+
  theme_classic()

home <- function(x) 500-0.5*x
graph+
  stat_function(fun = home, size=1.1)+
  scale_y_continuous(expand=c(0,0),limits=c(0,1000), breaks = c(0,500,1000))+
  scale_x_continuous(expand=c(0,0),limits=c(0,1000), breaks =c(0,500,1000))+
  labs(x="Food (kg)",y="Cloth (m)")
```
:::

## Opportunity cost

- The opportunity cost of making food is how many meter of cloth Home must stop producing in order to produce 1 kg of food:

$$
\frac{a_{LF}}{a_{LC}}
$$

- the opportunity cost is constant because the unit labor requirements are both constant.

- the opportunity cost of food appears as the absolute value of the slope of the PPF:

## PPF

$$
Q_F=\frac{L}{a_{LF}}-\left( \frac{a_{LF}}{a_{LC}} \right) Q_C
$$

- First difference, you get:

$$
\frac{dQ_F}{dQ_C}=-\frac{a_{LF}}{a_{LC}}, \text{in our case,}=-\frac{1}{2}
$$

## opportunity cost

For the case of $Q_F+2Q_C \leq 1000$:

to increase 1 kg of food, we need 1 hour of labor. But 1 hour of labor equals to 0.5 m of cloth (because we need 2 hours of labor to make 1 m of cloth).

Likewise, to have 1m more of cloth, we need 2 hour of labor, which could produce 2 kg of food instead.

| to produce | opportunity cost |
| --- | --- |
| 1 kg of food | 1/2 m of cloth |
| 1 m of cloth | 2 kg of food |

## Relative prices {.s}

- Workers will choose to work in an industry with higher wage.

- If the price of food relative to the price of cloth exceeds the opportunity cost of producing food:

$$
\frac{P_F}{P_C}>\frac{a_{LF}}{a_{LC}}
$$

- then the wage paid when making food will exceed the wage in cloth.

$$
W_F=\frac{P_F}{a_{LF}} > \frac{P_C}{a_{LC}}=W_C
$$

- so workers will only make food (specialize in food)

## Example

- Say $P_F=\$4$ while $P_C=\$7$.

- Then, $W_F=\frac{P_F}{a_{LF}}=\$4/hr$

- Also, $W_C=\frac{P_C}{a_{LC}}=\$3.5/hr$

- Relative wage of food, 4/7, is smaller than opportunity cost of food,1/2.

- Therefore, all workers will just make Food.

## Relative prices

- If home country wants to produce both, then wage in both industries have to be equal.

$$
W_F=W_C \Rightarrow \frac{P_F}{a_{LF}}=\frac{P_C}{a_{LC}}
$$

- Labor Theory of Value (LTV):

$$
\frac{P_F}{P_C}=\frac{a_{LF}}{a_{LC}}
$$

## Trade in Ricardian

- Introducing foreign country, notified with "*", with different technology.

> A country $j$ has an absolute advantage in producing good $i$ if that country can produce a unit of $i$ with less resources than another country

- If $a_{LF} < a^*_{LF}, home labor is more efficient in producing food than foreign because it needs less labor to make 1 kg of food.

- Does that guarantee that Home should export cheese?

## Trade in Ricardian

- What determines the pattern of trade is comparative advantage, not absolute advantage.

- Remember, comparative advantage is about the opportunity cost of production of good $i$

- If home country has a comparative advantage of Food:

$$
\frac{a_{LF}}{a_{LC}} < \frac{a^*_{LF}}{a_{LC}^*}
$$

## Trade in Ricardian

- Home exports food  if $\frac{a_{LF}}{a_{LC}} < \frac{a^*_{LF}}{a_{LC}^*}$

- Home exports cloth if $\frac{a_{LF}}{a_{LC}} > \frac{a^*_{LF}}{a_{LC}^*}$

- Both countries won't trade if $\frac{a_{LF}}{a_{LC}} = \frac{a^*_{LF}}{a_{LC}^*}$

- If home has absolute advantage in both goods but the ratio are the same with foreign, then home will have more stuff but they still won't trade.

## Foriegn's PPF

::: {.s layout-ncol="2"}

- if $\frac{a_{LF}}{a_{LC}} < \frac{a^*_{LF}}{a_{LC}^*}$, then foreign's ppf is steeper than home.

- That is, home sacrifice smaller number of cloth to make food than the opposite.

- In foreign, the reverse is true: making cloth requires less sacrifice of food than otherwise.

```{r}
#| fig-width: 5
#| fig-height: 4
#| label: fig2
#| fig-cap: "Production Possibility Frontier (PPF) country F"

foreign <- function(x) 1000-2*x
graph+
  stat_function(fun = foreign, size=1.1)+
  scale_y_continuous(expand=c(0,0),limits=c(0,1000), breaks = c(0,500,1000))+
  scale_x_continuous(expand=c(0,0),limits=c(0,1000), breaks =c(0,500,1000))+
  labs(x="Food (kg)",y="Cloth (m)")
```
:::

## Trade and prices

- Before trade, each countries' relative price of food reclects the relative price of food in terms of cloth.

- Therefore, relative price of food in Home is lower than relative price of food in Foreign.

- It will be profitable to ship food from Home to foreign (and cloth from foreign to home).

- What happens to price?

## Price after trade {.m}

- Let's work our way from low $\frac{P_F}{P_C}$ and increase it.

- $\frac{P_F}{P_C}<\frac{a_{LF}}{a_{LC}}<\frac{a_{LF}^*}{a_{LC}^*}$ no food is produced.

- $\frac{P_F}{P_C}=\frac{a_{LF}}{a_{LC}}<\frac{a_{LF}^*}{a_{LC}^*}$ Home indifferent, foreign produce only cloth.

- $\frac{a_{LF}}{a_{LC}}<\frac{P_F}{P_C}<\frac{a_{LF}^*}{a_{LC}^*}$ Home produce only food, Foreign produce only cloth.

- $\frac{a_{LF}}{a_{LC}}<\frac{P_F}{P_C}=\frac{a_{LF}^*}{a_{LC}^*}$ Home produce only food, Foreign indifferent

- $\frac{a_{LF}}{a_{LC}}<\frac{a_{LF}^*}{a_{LC}^*}<\frac{P_F}{P_C}$ no cloth is produced.

- Relative demand will help determine final relative price.

## Example

| a | Food | Cloth |
| --- | --- | --- |
| Home | $a_{LF}=1$ | $a_{LC}=2$ |
| Foreign | $a_{LF}=6$ | $a_{LC}=3$ |
| L | 1000 | 3000 |

::: {.incremental}

- Home has absolute advantages on both goods!

- What about comparative advantage?

- $\frac{1}{2}=\frac{a_{LF}}{a_{LC}}<\frac{a_{LF^*}}{a_{LC}^*}=2$

:::

## Example

| production | Home opportunity cost | Foreign Opportunity cost |
| --- | ----- | ----- |
| 1 kg of Food | 1/2 m of cloth | 2 m of cloth |
| 1 m of cloth | 2 kg of food | 1/2 kg of food |

- Home has a comparative advantage in producing food.

- Foreign has a comparative advantage in producing cloth.

## Prices

- Without trade, we know that $\frac{P_F}{P_C}$ are 1/2 in Home and 2 in Foreign.

- Suppose the global relative price $\frac{P_F}{P_C}=1$, means the world relative price of food is higher for Home, and lower in Foreign.

    - Home used to need to sacrifice 2 kg of food to get 1 m cloth. Now it's just 1.
    
    - Foreign used to need to sacrifice 2 m of cloth to get 1 kg of food. Now it's just 1.
    
## For Home

$$
Q_F+2Q_C \leq 1000
$$

- Say home used to demand 500 food and 250 cloth. Without trade, they achieve this by using L=500 to produce F and L=500 to produce C.

- With trade, they can use L=1000 to produce 1000 Food. Sell 500 Food for 500 cloth (because $\frac{P_F}{P_C}=1$), and ended up with 500 food and 500 cloth.

- Without trade, 500 food + 500 cloth would require them to have L=1500, which is unfeasible!

## Relative wage

- Our setting suggests Home to focus on producing Food while Foreign focuses on Cloth.

- Suppose that $P_F=12$ and $P_C=12$,

- Since Home produces Food, then $w_H=\frac{P_F}{a_{LF}}=\frac{12}{1}=12$

- Since Foreign produces Cloth, then $w_F=\frac{P_C}{a_{LC}}=\frac{12}{3}=4$

- Meaning, Home wage is 3 times Foreign (12/3).

## 

## PPF and Relative Prices

::: {.columns .s}
::: {.column width="48%"}
A simple labor theory of value:

> In the absence of international trade, the relative prices of goods are equal to their relative unit labor requirements.

-   if $P_A=P_B$, better concentrate on making A:
    -   wage $={1000\times P_A} \geq 500 \times P_B$
-   workers are better off making B only if $P_A < \frac{1}{2}P_B$
    -   workers only produce both iff $P_A = \frac{1}{2}P_B$
-   equilibrium is at $\frac{a_A}{a_B}=\frac{P_A}{P_B}$
:::

::: {.column width="48%"}
```{r}
#| fig-width: 5
#| fig-height: 4
#| label: fig3
#| fig-cap: "Production Possibility Frontier (PPF) negara H"
library('ggplot2')
graph<-ggplot()+
  theme_classic()

home <- function(x) 500-0.5*x
graph+
  stat_function(fun = home, size=1.1)+
  scale_y_continuous(expand=c(0,0),limits=c(0,1000), breaks = c(0,500,1000))+
  scale_x_continuous(expand=c(0,0),limits=c(0,1000), breaks =c(0,500,1000))+
  labs(x="Apel (kg)",y="Beras (kg)")
```
:::
:::

## PPF and Trade

-   To understand this term better, we will use some simplified hypothetical examples.

-   Suppose there are two economies, IDN and USA, and two goods, textiles (T) and soybean (S).

```{=tex}
\begin{equation} \label{eq1}
\begin{split}
IDN: \ 1T+4S&=200 \\
USA: \ 2T+2S & = 200 
\end{split}
\end{equation}
```
::: s
## 2 countries in autarky

```{=tex}
\begin{equation} \label{eq2}
\begin{split}
IDN: \ 1T+4S&=200 \\
USA: \ 2T+2S & = 200 
\end{split}
\end{equation}
```
| Produce           | US opportunity cost | IDN opportunity cost |
|-------------------|---------------------|----------------------|
| 1 ton of textiles | 1 ton of soybean    | 250 kg of soybean    |
| 1 ton of soybean  | 1 ton of textiles   | 4 ton of soybean     |

-   IDN is comparatively better at making textiles, while US is better at producing soybean:
    -   To make 1 ton of textiles, US needs to gave up resources which can be used to make 1 ton soybean, while Indonesians need only to give up 250 kg soybean.
    -   To make 1 ton of soybean, US needs to gave up resources which can be used to make 1 ton textiles, while Indonesians need only to give up 4 ton of textiles.
-   Note:

$$
\frac{a_{it}}{a_{is}}=\frac{1}{4} \geq \frac{1}{1}=\frac{a_{ut}}{a_{us}}
$$
:::

------------------------------------------------------------------------

Suppose the two countries do not trade. Indonesia's production is {T,S}={100,25}, While in the US, it's {50,50}.

```{r grafik1,echo=FALSE, message=FALSE, warning=FALSE}
#| fig-width: 7
#| fig-height: 5.5
#| label: fig4

library(patchwork)
library(kableExtra)
USA <- function(x) 100-x
IDN <- function(x) 50-.25*x

amrik <- graph +
  stat_function(fun = USA, color="blue", size=1.1)+
  scale_y_continuous(expand=c(0,0),limits=c(0,100), breaks = c(50,100))+
  scale_x_continuous(expand=c(0,0),limits=c(0,200), breaks =c(50,100))+
  labs(x="Quantity of textiles (ton)",y="Quantity of soybeans (ton)",
       title = "USA PPF",size=1)+
  annotate("label", x = 85, y = 70, label = 'US production \n and consumption \n in autarky' ,size=5)+
  geom_segment(aes(x = 50, y = 50, xend = 70, yend = 61)
               ,size=1)+
  geom_point(aes(x=50,y=50),color='black',size=4)+
  geom_segment(aes(x = 0, y = 50, xend = 50, yend = 50), lty = "dotted"
               ,size=1) +
  geom_segment(aes(x = 50, y = 0, xend = 50, yend = 50), lty = "dotted"
               ,size=1)
  
indon <- graph +
  stat_function(fun = IDN, color="red", size=1.1)+
  scale_y_continuous(expand=c(0,0),limits=c(0,100), breaks = c(50,25))+
  scale_x_continuous(expand=c(0,0),limits=c(0,200), breaks =c(100,200))+
  labs(x="Quantity of textiles (ton)",y="Quantity of soybeans (ton)",
       title = "IDN PPF",size=1)+
  annotate("label", x = 85, y = 60, label = 'IDN production \n and consumption \n in autarky' ,size=5)+
  geom_segment(aes(x = 100, y = 25, xend = 70, yend = 51)
               ,size=1)+
  geom_point(aes(x=100,y=25),color='black',size=4)+
  geom_segment(aes(x = 0, y = 25, xend = 100, yend = 25), lty = "dotted"
               ,size=2) +
  geom_segment(aes(x = 100, y = 0, xend = 100, yend = 25), lty = "dotted"
               ,size=2)

indon+amrik
```

## Gains from trade

```{r tab,echo=FALSE, message=FALSE, warning=FALSE}
game <- data.frame(
  "country"=c(rep('USA',times=2),rep('IDN',times=2)), "goods"=c("textiles (ton)","soybean (ton)","textiles (ton)","soybean (ton)"),
  "p1"=c(50,50,100,25), "p2"=c(50,50,100,25),"p3"=c(0,100,200,0),
  "p4"=c(75,62.5,125,37.5),"gains"=c("+25","+12.5","+25","+12.5")
)

zzz<-kbl(game, align = "c", format='html',
    col.names = c(" "," ", "production","consumption",
                     "production","consumption", "Gains from trade")) %>%
  kable_styling(font_size=18) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1) %>%
  add_header_above(c(" "," ", "in autarky" = 2,"with trade"=3))
zzz
```

By concentrating production on one good and then trade, you can see that both ended up better-off:

-   with autarky, total production {textiles,soybean} $=\{150,75\}$

-   consumption=production, and $C_{USA} = \{50,50\},C_{IDN} = \{100,25\}$

------------------------------------------------------------------------

## Gains from trade

```{r tab1,echo=FALSE, message=FALSE, warning=FALSE}
zzz
```

-   with trade, total production {textiles,soybean} $=\{200,100\}$

    -   Production $\Rightarrow Q_{USA}=\{0,100\}, Q_{IDN}=\{200,0\}$
    -   Consumption $\Rightarrow C_{USA}=\{75,62.5\}, C_{IDN}=\{125,37.5\}$

------------------------------------------------------------------------

```{r grafik2,echo=FALSE, message=FALSE, warning=FALSE,fig.width=10,fig.height=8}

amrik1 <- graph +
  stat_function(fun = USA, color="blue", size=1.1)+
  scale_y_continuous(expand=c(0,0),limits=c(0,100), 
                     breaks = c(50,62.5,100))+
  scale_x_continuous(expand=c(0,0),limits=c(0,200), 
                     breaks =c(50,75,100))+
  labs(x="Quantity of textiles (ton)",y="Quantity of soybeans (ton)",
       title = "USA PPF",size=0.5)+
  annotate("label", x = 130, y = 30, label = 'US production \n and consumption \n in autarky' ,size=5)+
  annotate("label", x = 65, y = 90, label = 'US production \n with trade' ,size=5)+
  annotate("label", x = 130, y = 62.5, label = 'US consumption \n with trade' ,size=5)+
  geom_segment(aes(x = 0, y = 100, xend = 35, yend = 90 )
               ,size=1)+
  geom_segment(aes(x = 75, y = 62.50, xend = 95, yend = 63)
               ,size=1)+
  geom_segment(aes(x = 50, y = 50, xend = 100, yend = 40)
               ,size=1)+
  geom_point(aes(x=50,y=50),color='black',size=4)+
  geom_point(aes(x=75,y=62.5),color='black',size=4)+
  geom_point(aes(x=0,y=100),color='black',size=4)+
  geom_segment(aes(x = 0, y = 50, xend = 50, yend = 50), lty = "dotted"
               ,size=1) +
  geom_segment(aes(x = 50, y = 0, xend = 50, yend = 50), lty = "dotted"
               ,size=1)+
  geom_segment(aes(x = 0, y = 62.5, xend = 75, yend = 62.5), lty = "dotted"
               ,size=1) +
  geom_segment(aes(x = 75, y = 0, xend = 75, yend = 62.5), lty = "dotted"
               ,size=1)

indon1 <- graph +
  stat_function(fun = IDN, color="red", size=1.1)+
  scale_y_continuous(expand=c(0,0),limits=c(0,100), breaks = c(50,37.5,25))+
  scale_x_continuous(expand=c(0,0),limits=c(0,200), breaks =c(100,125,200))+
  labs(x="Quantity of textiles (ton)",y="Quantity of soybeans (ton)",
       title = "IDN PPF",size=1)+
  annotate("label", x = 45, y = 80, 
           label = 'IDN production \n and consumption \n in autarky' ,
           size=5)+
  annotate("label", x = 95, y = 65,
           label = 'IDN consumption \n with trade' ,size=5)+
  annotate("label", x = 160, y = 30, 
           label = 'IDN production \n with trade' ,size=5)+
  geom_segment(aes(x = 100, y = 25, xend = 35, yend = 73)
               ,size=1)+
  geom_segment(aes(x = 125, y = 37.5, xend = 90, yend = 59)
               ,size=1)+
  geom_segment(aes(x = 200, y = 0, xend = 160, yend = 25)
               ,size=1)+
  geom_point(aes(x=100,y=25),color='black',size=4)+
  geom_point(aes(x=125,y=37.5),color='black',size=4)+
  geom_point(aes(x=200,y=0),color='black',size=4)+
  geom_segment(aes(x = 0, y = 25, xend = 100, yend = 25), lty = "dotted"
               ,size=1) +
  geom_segment(aes(x = 100, y = 0, xend = 100, yend = 25), lty = "dotted"
               ,size=1)+
  geom_segment(aes(x = 0, y = 37.5, xend = 125, yend = 37.5), lty = "dotted"
               ,size=1) +
  geom_segment(aes(x = 125, y = 0, xend = 125, yend = 37.5), lty = "dotted"
               ,size=1)
indon1+amrik1
```

------------------------------------------------------------------------

```{r}
#| fig-width: 5
#| fig-height: 4
#| label: fig5

graph<-ggplot()+
  theme_classic()
USA <- function(x) 100-x
IDN <- function(x) 50-.25*x
gab <- function(x) 100-.5*x
graph+
  stat_function(fun = gab, size=1.1)+
  stat_function(fun = USA, size=1.1,color="red")+
  stat_function(fun = IDN, size=1.1,color="blue")+
  scale_y_continuous(expand=c(0,0),limits=c(0,100), breaks = c(0,50,100))+
  scale_x_continuous(expand=c(0,0),limits=c(0,200), breaks =c(0,100,200))+
  labs(x="Quantity of Textiles (ton)",y="Quantity of Soybean (ton)")+
  geom_point(aes(x=100,y=25),color='black',size=4)+
  geom_point(aes(x=125,y=37.5),color='black',size=4)+
  geom_point(aes(x=75,y=62.5),color='black',size=4)+
  geom_point(aes(x=50,y=50),color='black',size=4)+
  annotate("label", x = 100, y = 30, label = 'CA[IDN]' ,size=5,parse=TRUE)+
  annotate("label", x = 125, y = 45, label = 'CT[IDN]' ,size=5,parse=TRUE)+
  annotate("label", x = 75, y = 70, label = 'CT[USA]' ,size=5,parse=TRUE)+
  annotate("label", x = 50, y = 55, label = 'CA[USA]' ,size=5,parse=TRUE)

```

------------------------------------------------------------------------

## Comparative advantage

-   In reality, we will need country's preferences to establish proper equilibrium quantity and prices.

-   In this model, we have no price. But from the opportunity cost, we can calculate the relative prices:

    -   In the US, opportunity cost to produce 1 soybean is 1 textile. That means, $P^{soybean}_{USA} = P^{textile}_{USA}$
    -   In IDN, opportunity cost to make 1 soybean is 4 textiles. That means, the price of soybean must be four times price of textile.

## Comparative advantage

-   Note that in autarky, both countries consume (and produce) with the same proportion as their unit labor cost.

-   With trade, both countries produce with 1:2 proportion

    -   their consumptions are skewed toward 1:2

-   This reflects the change in relative prices:

    -   with trade, textile gets relatively cheaper in US.
    -   in IDN, soybean gets a bit cheaper (from 1:4 to around 1:3.33)

------------------------------------------------------------------------

## Comparative VS Absolute advantage {.m}

-   It's possible that a country is good at producing many things.
    -   This is called absolute advantage.
-   There's no reason why US can't be better at making clothes than Indonesia.
    -   Hence, it's possible that in fact, US still better at making clothes
-   But what matters is *relative* opportunity cost:
    -   as long as its better for US capital and labour to produce soybean, they will still buy clothes from Indonesia.
    -   It may take lots of labor to produce textiles in Indonesia, but it'll take even more to produce soybean.
-   Makin specialised, makin besar opportunity cost.
    -   S2 UI vs SMEA.

------------------------------------------------------------------------

## Absolute advantage

-   Suppose USA dominates IDN in both:

```{=tex}
\begin{equation} \label{eq3}
\begin{split}
IDN: \ 2.5T+4S&=200 \\
USA: \ 2T+2S & = 200 
\end{split}
\end{equation}
```
-   Autarky: $Q_{USA}=\{50,50\}, Q_{IDN}=\{40,25\}$

-   Trade: $Q_{USA}=\{10,90\}, Q_{IDN}=\{80,0\}$

-   Total: $Q_{autarky}=\{90,75\}, Q_{trade}=\{90,90\}$

------------------------------------------------------------------------

## A note on absolute advantage

-   Mungkin anda menyadari bahwa absolute advantage hanyalah versi alternatif dari Walras Law.

-   Dengan kata lain, meski ada keunggulan absolut, *contract curve* & *better allocation* tetap ada (hence pareto improvement).

    -   Meski demikian, negara yang punya keunggulan absolut tetap memiliki posisi tawar yang lebih baik.

-   *trade is pareto improving* in the sense that we compare country A before trade with country A after trade.

    -   we do not compare country A with country B.

------------------------------------------------------------------------

## Taking from the poor?

-   In short, because IDNsians are less productive compared to the US, it's better to US to buy cheaper goods to IDN.

-   This is not to say that US take advantage from IDN:

    -   Since IDN labour is less productive anyway, they will still have low wage even if they sell domestically.
    -   In fact, serving US market may give them higher wage.

-   For example, if you feel ordering go-food is taking advantage from tukang ojeg, the alternative is even worse:

    -   if you ended up stop ordering go-food, their income is gone.

------------------------------------------------------------------------

## What determines comparative advantage? {.m}

1.  **Differences in climate** is the reason why we're so good at producing CPO and rubber, but sucks at producing soybean and wheat.

2.  **Differences in Factor Endowment**. Some countries are endowed with natural resource, some with cheap labour. Countries which has no both has to find something else, such as:

3.  **Differences in technology**. Japan, South Korea and Taiwan are one good example. While technology can be transferred, opportunity cost of investing in high-tech things is more production of CPOs.

------------------------------------------------------------------------

## Trade is Good {.m}

-   Barusan kita mempelajari bahwa ada gains dari trade bahkan ketika trading partner punya absolute advantage atas semua produksi.

-   Jika tidak ada pun, *autarky* akan terjadi tanpa dipasang penghalang / *barier*

    -   ini gak papa jika *autarky* terjadi tanpa pemaksaan.

-   Untuk analisis yang menyeluruh, kita perlu memperhitungkan *preference*

    -   *preference* dapat direfleksikan dengan harga.
    -   namun intuisinya tidak berbeda jauh.
    -   kita bahas di pertemuan-pertemuan berikutnya.

## Multi-goods comparative advantage

-   Misalnya ada N goods.

-   unit labor cost di home jadi $a_i$ di mana $i \in (1,N), N>2$

    -   correspondingly, $a_i^*$ adalah untuk foreign

-   untuk setiap good $i$, kita buat $\frac{a_i}{a_i^*}$

-   lalu urutkan sehingga $\frac{a_1}{a_1^*} < \frac{a_2}{a_2^*} < ... < \frac{a_N}{a_N^*}$

    -   alias diurutkan berdasarkan comparative advantage.

## Multi-goods comparative advantage

-   Let $w$ be home wage per OJ, while $w^*$ be foreign wage.
    -   artinya, biaya membuat 1 apel di home $= wa_{apel}$
    -   biaya membuat 1 apel di foreign $= w^*a_{apel}^*$
-   Goods are produced in the cheapest place.
    -   Apel dibikin di home jika: $wa_{apel}<w^*a_{apel}^*$
    -   atau jika: $\frac{a_{apel}^*}{a_{apel}}>\frac{w}{w^*}$

## Multi-goods comparative advantage {.m}

-   Let 2 economies with this characteristics:

| Good $(i)$ | $a_i$ | $a_i^*$ | $\frac{a_i^*}{a_i}$ |
|------------|:-----:|:-------:|:-------------------:|
| Apel       |   1   |   10    |         10          |
| Beras      |   5   |   40    |          8          |
| CPO        |   3   |   12    |          4          |
| Durian     |   6   |   12    |          2          |
| Emas       |  12   |    9    |        0.75         |

-   Note that table above is sorted decendingly by $\frac{a^*_i}{a_i}$

-   or: $\frac{a_1}{a_1^*} < \frac{a_2}{a_2^*} < ... < \frac{a_5}{a_5^*}$

## Multi-goods comparative advantage {.m}

| Good $(i)$ | $a_i$ | $a_i^*$ | $\frac{a_i^*}{a_i}$ |
|------------|:-----:|:-------:|:-------------------:|
| Apel       |   1   |   10    |         10          |
| Beras      |   5   |   40    |          8          |
| CPO        |   3   |   12    |          4          |
| Durian     |   6   |   12    |          2          |
| Emas       |  12   |    9    |        0.75         |

-   Ingat: dibikin di home jika: $\frac{a_{i}^*}{a_{i}}>\frac{w}{w^*}$

-   say $\frac{w}{w^*}=\frac{12}{4}=3$

-   Maka apel, beras dan CPO akan dibuat domestik.

-   Durian dan Emas diimpor.

## How useful?

-   Relative productivity dan relative wage merupakan poin analisis yang sangat intuitif.
    -   gets complicated very soon as we add countries and factors.
    -   tastes, economic structures, etc
    -   but the intuition is very similar.
-   Namun jika analisis ini benar, berarti negara-negara akan auto-spesialisasi? Apakah ini yang kita observe di dunia nyata?

## What prevents specialisation?

-   More than 1 factors
-   Protectionisms.
    -   more on this later.
-   Transportation cost.
    -   Some places are inherently difficult to reach.
    -   Some goods are harder to move.
-   at the extreme, a good can be **nontradable good**.
    -   most services are like this.

------------------------------------------------------------------------

## What prevents specialisation?

-   Ricardian model: 1 very mobile factor of production.

-   In reality, trade matters in distribution of income.

    -   short-run: moving factors are costly.
    -   long-run: shift in factor demand.

-   Factor owners (land, labor, capital, etc) receive different gains from trade!

## Japanese rice farmer case:

-   Land scarcity -\> rice is expensive (even compared to US).
-   Free trade means higher real wage.
-   Farmers may need to move industries.
    -   reskilling is costly.
    -   land value drops.
    -   machines & infrastructure become useless.
-   Better spend money to organize politically!

## Specific factors

-   Next, we add one more factor to the model: a specific factor.

-   Labor can move between sectors, but this one other factor is specific to the sector, hence can't move freely.

    -   just like our Japanese farmers!

-   We will also discuss a bit what is this specific factor of production in reality.

------------------------------------------------------------------------

::: r-stack
<br/>

<br/>

<iframe src="https://giphy.com/embed/3ohs86vZAWiJXWvQI0" width="480" height="353" frameBorder="0" class="giphy-embed" allowFullScreen>

</iframe>

<p><a href="https://giphy.com/gifs/eminem-my-name-is-the-slim-shady-lp-3ohs86vZAWiJXWvQI0">via GIPHY</a></p>
:::
