---
title: "Argument against free trade"
subtitle: "ECES905205 pertemuan 6"
date: "3 October 2023"
author: "I Made Krisna Gupta"
format: 
  revealjs:
    theme: [simple,../mytheme.scss]
    chalkboard: true
    incremental: false
    transition: convex
    background-transition: fade
    slide-number: true
  beamer: default
---

## Today

- We have learned why trade is good.

- These models however uses various assumptions that may or may not be true.

- When these assumptions are relaxed, some reason to NOT trade may emerges.

- Today we learn two: small country & dynamic efficiency.

- Leads to 3 insights: optimal tariff, infant industry, and current account management.

## Gains from trade recap

- So far, we have explored gains from trade:

    - difference in technology $\rightarrow$ specialization (comparative advantage)
    
    - less constraint (i.e., as long as trade is balanced)
    
    - difference in factor endowment
    
- With large [fixed cost & monopolistic]{.underline}, we get loves of variety & internal economies of scale efficiency.

- Remember, as long as we can reach higher bundle with trade, then trade is better!

## Argument agains trade

- Gains of trade is not shared equally if [factor movements and ownerships cannot switch fluidly]{.underline}.

    - factor owners that stay in the import-competing industry loses.
    
    - violates pareto optimality.
    
- We showed that in this case, overall economic size increases, so in theory we can compensate those losers.

- First policy lesson: we can redistribute losers as long as the pie is larger (and the government is efficient)

## Large country assumption

- We use small country assumption so far.

- That is, whatever happens in our country in question, it will not affect the world market.

- In reality, various countries' action affects world prices!

    - OPEC capacity reduction increases oil price. US entered a recession because of this.
    
    - Indonesian nickel ban $\rightarrow$ increased nickel ore price.
    
## Large country assumption

- Remember, in international trade, we often use ToT (i.e., $\frac{P_X}{P_M}$) to measure welfare.

- Since large countries can affect prices while its small partners cannot, large countries can alter their policy to affect ToT in their favor.

- Basically kinda like a monopolist.

- To see the mechanism, we have to understand the impact of tariff for small country first.

## Tariff for small country

```{r}
#| label: fig1
#| warning: false
library(WDI)
library(tidyverse)
library(patchwork)

D<-function(q) 10-q
S<-function(q) q

a<-ggplot()+
  scale_x_continuous(expand=c(0,0),limits = c(0,10),breaks=c(3,4,6,7),
                     labels=c("Sm","St","Dt","Dm"))+
  scale_y_continuous(expand=c(0,0),limits = c(0,10),breaks=c(3,4),
                     labels=c("Pw","Pw+t"))+
  geom_function(fun = S,size=1, color='red') +
  stat_function(fun = D,size=1, color='blue')+
  annotate("segment",x=0,xend=6,y=4,yend=4,lty="dashed")+
  annotate("segment",x=0,xend=7,y=3,yend=3,lty="dashed")+
  annotate("segment",x=4,xend=4,y=0,yend=4,lty="dashed")+
  annotate("segment",x=6,xend=6,y=0,yend=4,lty="dashed")+
  annotate("segment",x=7,xend=7,y=0,yend=3,lty="dashed")+
  annotate("segment",x=3,xend=3,y=0,yend=3,lty="dashed")+
  annotate("text",x=2,y=3.5,label="A")+
  annotate("text",x=3.7,y=3.2,label="B")+
  annotate("text",x=5,y=3.5,label="C")+
  annotate("text",x=6.2,y=3.2,label="D")+
  labs(x="Q",y="P")+
  theme_classic()+
  theme(axis.title.x = element_text(hjust=1),
        panel.background = element_rect(fill = "#f0f1eb", colour = "#f0f1eb"),
        plot.background = element_rect(fill = "#f0f1eb"),
        axis.title.y = element_text(angle=0,hjust=1,vjust=1),
        panel.grid=element_blank(),
#        axis.text.x=element_blank(),
#        axis.text.y=element_blank(),
#        axis.ticks.x=element_blank(),
#        axis.ticks.y=element_blank(),
        legend.position=c(1,1),
        legend.justification = c(1,1),
        legend.spacing=unit(0,"cm"),
        legend.margin=margin(0,0,0,0))
a
```

## Welfare impact

- Import lowers from $D_m - S_m$ to $D_t-S_t$
- A+B+C+D = CS loss.
- A = CS that transfer to PS.
- C = tax revenue = CS that transfers to the government.
- B+D = Deadweight Loss.

## Importer is a large country

```{r}
#| label: fig2
#| warning: false

demand<-function(q) 4-.7*q
supply<-function(q) .7*q
demand1<-function(q) 7.5-1.5*q
supply1<-function(q) 1.5*q

a<-ggplot()+scale_x_continuous(limits = c(1,5))+
  scale_y_continuous(limits = c(1,5))+theme_classic()+
  theme(axis.title.x = element_text(hjust=1),
        axis.title.y = element_text(angle=0,hjust=1,vjust=1),
        panel.grid=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.x=element_blank(),
        axis.ticks.y=element_blank(),
        legend.position=c(1,1),
        legend.justification = c(1,1),
        legend.spacing=unit(0,"cm"),
        legend.margin=margin(0,0,0,0))
b<-a+
  labs(x=expression(Q[H]),y=expression(P[H]),color=NULL,fill=NULL)+
  geom_function(fun = supply,size=1, color='red') +
  stat_function(fun = demand,size=1, color='blue')
c<-a+
  labs(x=expression(Q[F]),y=expression(P[F]),color=NULL,fill=NULL)+
  geom_function(fun = supply1,size=1, color='red') +
  stat_function(fun = demand1,size=1, color='blue')
c+a+b
  
```

## Importer is a large country

- Import tariffs shift world price.
  - $P_t - P_{tt}$ is the tariff rate.
  - There's ToT gain cuz $P_M$ goes down.
- US tariffs always shakes the world. Also when China used tariff to punish Australia when Australia endorsed investigation on the origin of SARSCOV2.

## Importer is a large country

```{r}

a<-ggplot()+
  scale_x_continuous(expand=c(0,0),limits = c(0,10),breaks=c(3,4,6,7),
                     labels=c("Sm","St","Dt","Dm"))+
  scale_y_continuous(expand=c(0,0),limits = c(0,10),breaks=c(2,3,4),
                     labels=c("Ptt","Pw","Pt"))+
  geom_function(fun = S,size=1, color='red') +
  stat_function(fun = D,size=1, color='blue')+
  annotate("segment",x=0,xend=6,y=4,yend=4,lty="dashed")+
  annotate("segment",x=0,xend=7,y=3,yend=3,lty="dashed")+
  annotate("segment",x=4,xend=4,y=0,yend=4,lty="dashed")+
  annotate("segment",x=6,xend=6,y=0,yend=4,lty="dashed")+
  annotate("segment",x=7,xend=7,y=0,yend=3,lty="dashed")+
  annotate("segment",x=3,xend=3,y=0,yend=3,lty="dashed")+
  annotate("segment",x=0,xend=7,y=2,yend=2,lty="dashed")+
  annotate("text",x=2,y=3.5,label="A")+
  annotate("text",x=3.7,y=3.2,label="B")+
  annotate("text",x=5,y=3.5,label="C")+
  annotate("text",x=6.2,y=3.2,label="D")+
  annotate("text",x=5,y=2.5,label="E")+
  labs(x="Q",y="P")+
  theme_classic()+
  theme(axis.title.x = element_text(hjust=1),
        panel.background = element_rect(fill = "#f0f1eb", colour = "#f0f1eb"),
        plot.background = element_rect(fill = "#f0f1eb"),
        axis.title.y = element_text(angle=0,hjust=1,vjust=1),
        panel.grid=element_blank(),
        legend.position=c(1,1),
        legend.justification = c(1,1),
        legend.spacing=unit(0,"cm"),
        legend.margin=margin(0,0,0,0))
a
```

## Welfare impact of large country's tariff

- DifferenceL import price changes $P_w \rightarrow P_{tt}$, hence A,B,C and D gets smaller.
- Government revenue = C+E
  - C from CS
  - E from TOT gain.
- if B+D < E, then there's a welfare gain from the tariff.
- This tariff hurts small exporters. Globally, it reduces welfare.

## Tariff and retaliation

- Bottom line: for a large country, they can gain from import tariff (if big in terms of buyer) or export tariff (if big in terms of exporters)

- If their partner is small, whatever they do won't affect world price.

- However, if their partner is large (or a combination of smaller countries like EU), then they can retaliate with their own tariff.

- Large partners' retaliation would reduce large country's $P_X$ which led to initial ToT.

## Free trade agreement

- Large countries can only benefit from tariffs if:
    
    - they are large enough (ensure B+D < E)
    
    - Didn't get retaliated a similarly sized partner(s).

- The second part leads to a prisoner's dilemma situation.

- WTO and trade agreements exist to tackle this situation.

- When a large country slap a unjustified tariff, small country can sue and other small countries can follow, combined creates large country effect.

## Dynamic efficiency

- Our model so far is static: cloth producer will always be a cloth producer.

- In reality, a country's industry structure changes: the Asian Tigers started as an agriculture economy.

- Some firms may ask for a **short-term** tariff protection to build scale and expertise before they are able to compete against foreign producers.

- That is, protection will lead to a short-term efficiency loss, but the promised long-run gain will make up for it.

- This is called **infant industry** argument.

## Infant industry

- In the infant industry case, a case for tariff protection isn't to avoid losses from import competition, but to create time for domestic industry.

- Protection will lead to markups (which is bad for consumers), which then is used to innovate and grow.

- Protection can be in the form of various policies: Tariffs, quota restrictions, Local Content Requirements (LCR), etc.

- In designing policy with infant industry as its main argument, two things must be made clear:

    - Measurement of the LEARNING of the industry & WHEN the policy will be revoked.

## Why not infant industry?

- If an industry has a potency in the future, why capital market won't finance their learning?

    - Tech startups do not need the government: they simple get investors to fund their learning.
    
- Instead of trade policy, why not use subsidy or fiscal incentives?

    - trade policy looks free, but it is paid by the consumers through inefficiency / DWL.
    
    - Non-tariff is even worse since its complicated and gains are unclear.
    
    - Also, retaliations -> more loss, less scale.
    
## Why not infant industry?

- What if the industry never grow up?

- Instead of growing, more firms actually just consume the markups.

    - SOEs, Steel, Automotive in Indonesia still not competitive.
    
- What if international firms are also growing?

- How to make sure the protected domestic firms can catch up? When a government should cut its losses?

## Import substitution strategy

- Since infant industry mostly argued for import competing industry, it often dubbed **Import Substitution Strategy** (ISS).

- Protection against good that a country is not comparatively good at will result in higher cost.

    - India & Pakistan protects heavy-manufacturing goods (cars, steel), but their main exports are light-manufacturing goods (textiles)
    
- Needs to ensure those industry exports BECAUSE of past protection.

## ISI vs Export orientation

- Choosing ISI = choosing not to pursue export orientation growth:

    - protection skews up prices for importing industries relative to exporting industries (e.g., $\frac{P_X}{P_M} \Downarrow$).
    
    - Factor of production will move to these industries instead of exporting industries.
    
- Again, this is bad for welfare, unless the ISI industries can catch up.

## ISI

- ISI most often ended up accompanied by increased FDI:

    - Incoming FDI seeks market.
    
    - This means, the domestic markups can be absorb by foreigners.
    
- However, domestic firms can learn from foreigners.
    
- Good examples are the rise of local champions: Toyota, Samsung, BYD, Foxconn.

    - What's Indonesia's local champions?
    
## ISI failure

- Often bad for welfare amid naturally pursue a comparative disadvantage industry.

- The forever infants: why grow if can reap markups?

- Markups absorbed by foreigners.

- Retaliations can be bad and actually reduce scale.

- Hard to measure and evaluate.

## Industrial policy

- If an industry has the potency to grow, theory says that a good capital market will comes in to finance that industry.

- However, if this is not the case, then ISI may be useful.

- In the presence of market failures, industrial policy that support ISI may be desirable:

    - Imperfect capital market
    
    - Coordination problem
    
- Next week, we will discuss the new industrial policy from these exact problems.