---
title: "Ilmu Ekonomi"
subtitle: "Pertemuan 4"
author: "Prodi PIWAR Politeknik APP Jakarta"
format: 
  revealjs:
    theme: [simple,../mytheme.scss]
    chalkboard: true
    incremental: false
    transition: convex
    background-transition: fade
    slide-number: true
  pptx:
    prefer-html: true

---

## Recap on last week

- We use consumer and producer surplus to measure the efficiency of the economy. We want the combination of both to be generally bigger.

- The two face a constant political battle: consumer wants small prices, producers like high prices.

- When market is perfect, intervention often leads to various types of inefficiency

## Today

- We will learn about one more intervention: tax.

- before we go there, it is best to understand elasticity first.

## What's an elasticity?

- If you are a business owner, understanding elasticity is important aspect to get maximized profit.

- **Price elasticity of demand** of a good is a concept of how much the demand of that good change  its price changes.
  - we usually just say **elasticity** in short.
  
- Maximizing profit requires this knowledge:
  - sometimes it is better to sell cheap to gain huge market share, or sell high if niche.
  
## Calculating elasticity

:::: columns
::: {.column width=47%}
- Consider the demand curve on the right.

- At $P_1=10$, the daily traffic is $Q_1=140$

- When $P_2=11$, daily traffic drops to $Q_2=127.3$
:::

::: {.column width=47%}
```{r grafik1,echo=FALSE, message=FALSE, warning=FALSE}
#| fig-width: 10
#| fig-height: 8
library(dplyr)
library(ggplot2)
library(ggdark)

demand <- function(q) (q/1400)^(-1)
q<-c(127.3,140)
p<-c(11,10)

graph<-ggplot()+
  dark_theme_bw(base_size=25)+
  theme(text = element_text(color='#e0e0e0'),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "#e0e0e0"))

grafik1 <- graph +
  stat_function(fun = demand,size=1.1, color='cyan') +
  scale_x_continuous(limits=c(120,150), 
                     breaks = q) +
  scale_y_continuous(limits=c(9,12), 
                     breaks = p) +
  geom_segment(data = data_frame(q=127.3,p=11),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=127.3,p=11),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=140,p=10),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=140,p=10),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2)+
  labs(title="Toll road demand \n for Surabaya-Gempol", x="Q (1000 cars/day)", y="P (1000 IDR)")+
  geom_text(aes(x = 150, y = 9.7, label = "D"), parse = TRUE, size =8, color = '#e0e0e0')
grafik1

```
:::
::::

## Calculating elasticity {.s}

- To calculate elasticity of demand, we first calculate:
  1. percent change in the quantity demanded
  1. percent change in the price

- Formula for no.1:

$$
\tag{1}
\text{% change in }Q^D=\frac{Q_2^D-Q_1^D}{Q_1^D}\times 100
$$


- Formula for no.2:

$$
\tag{2}
\text{% change in P}=\frac{P_2-P_1}{P_1}\times 100
$$


## Calculating elasticity {.m}

- for our case, we have:

$$
\begin{align*}
Q_1^D&=140 & Q_2^D&=127.3 \\ P_1&=10 & P_2&=11
\end{align*}
$$

- Thus:

$$
\begin{align}
\text{% change in }Q^D &= \frac{127.3-140}{127.3}\times 100 \\ &=-10 \%
\end{align}
$$

$$
\begin{align}
\text{% change in P} &= \frac{11-10}{10}\times 100 \\ &=10\%
\end{align}
$$

## Calculating elasticity {.s}

- After we get the numbers from {1} and {2}, we use this formula:

$$
\tag{3}
\text{Price elasticity of demand}=\frac{\text{% change in }Q^D}{\text{% change in P}}
$$


- in our case, we have:

$$
\begin{align}
\text{Price elasticity of demand}&=\frac{-10\%}{10\%} \\ &=-1
\end{align}
$$

## Calculating elasticity

- Economists are also lazy in writing the minus sign.

- Usually we report demand elasticity in **absolute term**.
  - i.e., we drop the minus sign.
  
- However, since most demand curves are downward sloping, it must be negative
  - hence we automatically knows what it means even when we report in absolute term.

## About percent change

Using percent changes is more useful than absolute changes
- Things are usually move on proportion
  - A wealthy person will feel casual getting 1 million IDR because it's probably just a small % of her total wealth.
    
- Quantity and prices have many metrics
  - kgs, litres, units, pairs, units/day, etc
  - IDR, USD, btc, etc.

## interpreting elasticity

<iframe src="https://giphy.com/embed/WYEWpk4lRPDq0" width="400" height="400" frameBorder="0" class="giphy-embed" allowFullScreen></iframe><p><a href="https://giphy.com/gifs/cat-WYEWpk4lRPDq0">via GIPHY</a></p>

## Interpreting elasticity

- We have seen the elasticity of our toll road case is around 1, but what does this mean?

- Note that the elasticity formula has % change in Q at the top, and % change in P at the bottom.

- In a sense, the number shows how much quantity change given the price change.

- that means, the bigger the number of the price elasticity, **the more elastic it is**

## A market is elastic. So what?

Let's have a look at different 'kind' of elasticities and what they mean

We'll begin with two extreme cases of elasticity: zero (0) and infinity ( $\infty$ )

## Two extreme cases

- Suppose the toll road is the only way to go to work.
  - No matter how high the price, people will still need to use it.
  - Hence, change in the demand of toll road is zero for any change in price.
  
- Suppose shuttlecocks are priced 10.000 IDR
  - If people don't care about the shuttlecocks' color, everyone will buy the pink one if the price drop to 9.900
  - Everyone won't buy a single one if the price go up only to 10.100 IDR

## Two extreme cases of elasticity

:::: columns
::: {.column width=47%}
```{r grafik2, message=FALSE, warning=FALSE, echo=FALSE}
grafik2 <- graph +
  geom_vline(xintercept=125, color = "cyan",size=2)+
  scale_x_continuous(limits=c(120,130), 
                     breaks = 125) +
  scale_y_continuous(limits=c(9,12), 
                     breaks = p) +
  geom_segment(data = data_frame(q=125,p=11),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=125,p=10),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2)+
  labs(title="Perfectly inelastic demand (0)", x="Q", y="P")+
  geom_text(aes(x = 125.5, y = 9, label = "D[1]"), parse = TRUE, size =8, color = '#e0e0e0')
grafik2
```
an illustration of **perfectly inelastic demand**
:::

::: {.column width=47%}
```{r grafik3, message=FALSE, warning=FALSE, echo=FALSE}
library(latex2exp)
grafik3 <- graph +
  geom_hline(yintercept=10.5, color = "cyan",size=2)+
  scale_x_continuous(limits=c(120,130), 
                     breaks = c(123,127)) +
  scale_y_continuous(limits=c(9,12), 
                     breaks = 10.5) +
  geom_segment(data = data_frame(q=123,p=10.5),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=127,p=10.5),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2)+
  labs(title="Perfectly Elastic Demand (\U221E)", x="Q", y="P")+
  geom_text(aes(x = 130, y = 10.7, label = "D[2]"), parse = TRUE, size =8, color = '#e0e0e0')
grafik3
```
an illustration of **perfectly elastic demand**
:::
::::

## Types of elasticities

- Most goods are somewhere in between.

- We usually use elasticity=1 as another important division point.

- 1 is a useful barrier because when elasticity=1, that means a change in price by 1 % leads to a change in quantity demanded by 1 %

- elasticity < 1 means the quantity response less than a change in prices, while elasticity > 1 means the quantity is highly responsive, more than the change in price.

---

:::: columns
::: {.column width=47%}
```{r grafik4,echo=FALSE, message=FALSE, warning=FALSE}

demand1 <- function(q) (q/433)^(-1/0.5)
demand2 <- function(q) (q/7250)^(-1/1.7)

grafik4 <- graph +
  stat_function(fun = demand1,size=1.1, color='cyan') +
  scale_x_continuous(limits=c(120,150), 
                     breaks = c(130.5,137)) +
  scale_y_continuous(limits=c(9,12), 
                     breaks = p) +
  geom_segment(data = data_frame(q=130.5,p=11),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130.5,p=11),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=137,p=10),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=137,p=10),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2)+
  labs(title="Inelastic demand (<1)", x="Q (1000 cars/day)", y="P (1000 IDR)")+
  geom_text(aes(x = 145, y = 9.3, label = "D"), parse = TRUE, size =8, color = '#e0e0e0')

grafik5 <- graph +
  stat_function(fun = demand2,size=1.1, color='cyan') +
  scale_x_continuous(limits=c(120,150), 
                     breaks = c(123,144.6)) +
  scale_y_continuous(limits=c(9,12), 
                     breaks = p) +
  geom_segment(data = data_frame(q=123,p=11),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=123,p=11),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=144.6,p=10),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=144.6,p=10),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2)+
  labs(title="Elastic demand (>1)", x="Q (1000 cars/day)", y="P (1000 IDR)")+
  geom_text(aes(x = 150, y = 10, label = "D"), parse = TRUE, size =8, color = '#e0e0e0')

grafik4
grafik5

```
:::
::: {.column width=47%}

```{r grafik5,echo=FALSE, message=FALSE, warning=FALSE}
grafik1 + labs(title="Unit elastic demand (=1)", x="Q (1000 cars/day)", y="P (1000 IDR)")

```
</br>
</br>
</br>

The flatter the curve, the closer its elasticity to $\infty$

:::
::::

## Types of demand elasticities {.m}

| Elasticity | meaning |
| ---- | ------- |
| 0 | **perfectly inelastic**. Price doesn't matter |
| < 1 | **inelastic**. Quantity is less responsive to price changes |
| 1 | **unit elastic**. Quantity change exactly the same as the price change in % terms|
| >1 | **elastic**. Quantity response heavily to price changes |
| $\infty$ | **perfectly elastic**. A slight increase of price reduces demand to zero |

## Elasticities vary. So what?

- Suppose you wan't to increase the revenue from the toll road operation, what price will you set?

- It is intuitive to think that the higher the price, the better.
  - High price $\Rightarrow$ more revenue

- Is there a problem with this logic?

- Remember, demand respond to price. High price $\Rightarrow$ lower demand $\Rightarrow$ less revenue

- The maximum price depends on the demand elasticity

---

```{r grafik6,echo=FALSE, message=FALSE, warning=FALSE}
grafik6 <- graph +
  stat_function(fun = demand1,size=1.1, color='cyan') +
  scale_x_continuous(limits=c(120,150), 
                     breaks = 137) +
  scale_y_continuous(limits=c(9,12), 
                     breaks = 10) +
  geom_segment(data = data_frame(q=137,p=10),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=137,p=10),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2)+
  geom_ribbon(aes(x=seq(0,137,1), ymin=9, ymax=10),alpha = 0.4,fill='red')+
  annotate("text", x = 128, y = 9.8, label = "Total revenue=", size=8)+
  annotate("text", x = 128, y = 9.6, label = "Price x Quantity", size=8)+
  annotate("text", x = 128, y = 9.4, label = "10 x 137", size=8)+
  annotate("text", x = 128, y = 9.2, label = "1370 mil IDR/day", size=8)+
  labs(title="Inelastic demand (<1)", x="Q (1000 cars/day)", y="P (1000 IDR)")+
  geom_text(aes(x = 145, y = 9.3, label = "D"), parse = TRUE, size =8, color = '#e0e0e0')
grafik6
```

---

```{r grafik7,echo=FALSE, message=FALSE, warning=FALSE}
grafik7 <- graph +
  stat_function(fun = demand1,size=1.1, color='cyan') +
  scale_x_continuous(limits=c(120,150), 
                     breaks = 130.5) +
  scale_y_continuous(limits=c(9,12), 
                     breaks = 11) +
  geom_segment(data = data_frame(q=130.5,p=11),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130.5,p=11),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2) +
  geom_ribbon(aes(x=seq(0,130.5,1), ymin=9, ymax=11),alpha = 0.4,fill='red')+
  annotate("text", x = 125, y = 10.4, label = "Total revenue=", size=7)+
  annotate("text", x = 125, y = 10.2, label = "P x Q", size=8)+
  annotate("text", x = 125, y = 10, label = "11 x 130.5", size=8)+
  annotate("text", x = 125, y = 9.8, label = "1435.5", size=8)+
  labs(title="Inelastic demand (<1)", x="Q (1000 cars/day)", y="P (1000 IDR)")+
  geom_text(aes(x = 145, y = 9.3, label = "D"), parse = TRUE, size =8, color = '#e0e0e0')
grafik7
```

---

## Impact of price increase {.m}

:::: columns
::: {.column width=47%}
```{r grafik8,echo=FALSE, message=FALSE, warning=FALSE}
#| fig-width: 6
#| fig-height: 7
grafik4 +
  geom_ribbon(aes(x=seq(0,130.5,1), ymin=9, ymax=10, fill="B"),alpha = 0.4,show.legend=FALSE)+
  geom_ribbon(aes(x=seq(130.5,137,1), ymin=9, ymax=10, fill="A"),alpha = 0.4,show.legend=FALSE)+
  geom_ribbon(aes(x=seq(0,130.5,1), ymin=10, ymax=11, fill="C"),alpha = 0.4,show.legend=FALSE)+
  annotate("text", x = 125, y = 9.5, label = "B", size=9)+
  annotate("text", x = 134, y = 9.5, label = "A", size=9)+
  annotate("text", x = 125, y = 10.5, label = "C", size=9)
  
```
:::
::: {.column width=47%}

- Segment A is the **Quantity effect of price increase**
  - Negative effect since fewer units sold

- Segment C is the **Price effect of price increase**
  - positive effect since higher price for each unit sold
  
:::
::::

## Impact of price increase

:::: columns
::: {.column width=47%}
```{r grafik9,echo=FALSE, message=FALSE, warning=FALSE}
#| fig-width: 6
#| fig-height: 7
grafik4 +
  geom_ribbon(aes(x=seq(0,130.5,1), ymin=9, ymax=10, fill="B"),alpha = 0.4,show.legend=FALSE)+
  geom_ribbon(aes(x=seq(130.5,137,1), ymin=9, ymax=10, fill="A"),alpha = 0.4,show.legend=FALSE)+
  geom_ribbon(aes(x=seq(0,130.5,1), ymin=10, ymax=11, fill="C"),alpha = 0.4,show.legend=FALSE)+
  annotate("text", x = 125, y = 9.5, label = "B", size=9)+
  annotate("text", x = 134, y = 9.5, label = "A", size=9)+
  annotate("text", x = 125, y = 10.5, label = "C", size=9)
  
```
:::

::: {.column width=47%}
- If you are profit oriented, Increasing the price is a good idea if:

$$
B+C > B+A
$$

$$
C>A
$$
:::
::::

## Impact of price increase

:::: columns
::: {.column width=47%}
```{r grafik10,echo=FALSE, message=FALSE, warning=FALSE}
#| fig-width: 6
#| fig-height: 7
grafik5 +
  geom_ribbon(aes(x=seq(0,123,1), ymin=9, ymax=10, fill="B"),alpha = 0.4,show.legend=FALSE)+
  geom_ribbon(aes(x=seq(123,144.6,1), ymin=9, ymax=10, fill="A"),alpha = 0.4,show.legend=FALSE)+
  geom_ribbon(aes(x=seq(0,123,1), ymin=10, ymax=11, fill="C"),alpha = 0.4,show.legend=FALSE)+
  annotate("text", x = 122, y = 9.5, label = "B", size=9)+
  annotate("text", x = 134, y = 9.5, label = "A", size=9)+
  annotate("text", x = 122, y = 10.5, label = "C", size=9)
  
```
:::

::: {.column width=47%}
- In a market with high demand elasticity, the quantity effect outweighs the price effect.

- Increasing the price is not so clever if the market's demand is highly elastic.
:::
::::

## Last point on demand elasticity

- When elasticity = 1, then $A=C$
  - hence changing price has no impact on sales revenue.

- Profit oriented in the real life may not very realistic
  - Governments may want to serve as much people as possible
  - In the short run, growth and market cap may be more important than profit.

## Last point on demand elasticity

- We usually tied elasticity on a price point.
  - on our toll road case, our price point is 10.000 IDR
  
- Elasticity may differ in different price point.
  - Changing price from 10 to 11 may have different impact compared to changing price from 20 to 21.
      - the first one is an increase of 10%.
      - the second one is an increase of 5%.
      - in absolute term, both are a 1.000 IDR increase.

## Last point on demand elasticity

- Estimating elasticity is very hard

- There are many things also at play:
  - different condition in different time
  - different buyer with different characteristics behave differently
  - Income, taste, etc matters (remember what **shifts the demand curve?**)

## Factors determine demand elasticity

- A **necessity** good tend to have lower elasticity compared to a **luxury good**.

- The availability of **close substitutes**, like the case with pertamax and shell.

- **Share of income spent** on the good. Rich people may not care with toll road price.

- **Time**. Cigarettes may have low elasticity in the short run because it is so addictive, and adapt later in the long run.

## Other types of demand elasticity {.m}

- **Cross price elasticity of demand** is how the price of good $B$ affects a demand of good $A$.

$$
\tag{4}
\text{Cross-Price elasticity}_{A,B}=\frac{\text{% change in }Q_A}{\text{% change in }P_B}
$$


- When good A and good B are subtitutes, then the effect is negative.
  - e.g., Samsung and Xiaomi

- When good A and good B are complements, the effect is positive.
  - e.g., Samsung and Telkomsel

## Other types of demand elasticity {.m}

- **Income elasticity of demand** is how the chanage on income affects your demand of good $A$


$$
\tag{5}
\text{Income elasticity of Demand}_A=\frac{\text{% change in }Q_A}{\text{% change in Income}}
$$


- It is usually positive if good $A$ is a **normal good**.

- It is negative if good $A$ is an **inferior good**.
  - you buy less instant noodle when you're richer.
  
## Other types of demand elasticity {.m}

- **Income elasticity of demand** is how the chanage on income affects your demand of good $A$

$$
\tag{5}
\text{Income elasticity of Demand}_A=\frac{\text{% change in }Q_A}{\text{% change in Income}}
$$


- A **necessity** good is usually **income inelastic** because you need to buy it no matter what.

- A **luxury** good can be **income elastic**. You only buy it if your main needs are covered so you can wait for a discount.

# Price elasticity of Supply

## Price elasticity of Supply {.m}

- Supply also have elasticities. As a supplier, you can't afford to sell expensive if there are high competition.

- However, in many cases (such as toll road), entering the market is very hard.

- How easy it is to expand (or for new supplier to enter the market) matters to how responsive can supply react to the change in price.

$$
\tag{6}
\text{Price elasticity of supply} = \frac{\text{% change in }Q^s}{\text{% change in }P}
$$


## Two extreme cases of elasticity

:::: columns
::: {.column width=47%}
```{r grafik11, message=FALSE, warning=FALSE, echo=FALSE}
grafik11 <- graph +
  geom_vline(xintercept=125, color = "red",size=2)+
  scale_x_continuous(limits=c(120,130), 
                     breaks = 125) +
  scale_y_continuous(limits=c(9,12), 
                     breaks = p) +
  geom_segment(data = data_frame(q=125,p=11),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=125,p=10),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2)+
  labs(title="Perfectly inelastic supply (0)", x="Q", y="P")+
  geom_text(aes(x = 125.5, y = 9, label = "S[1]"), parse = TRUE, size =8, color = '#e0e0e0')
grafik11
```
.s[an illustration of **perfectly inelastic supply**]
:::

::: {.column width=47%}
```{r grafik12, message=FALSE, warning=FALSE, echo=FALSE}

grafik12 <- graph +
  geom_hline(yintercept=10.5, color = "red",size=2)+
  scale_x_continuous(limits=c(120,130), 
                     breaks = c(123,127)) +
  scale_y_continuous(limits=c(9,12), 
                     breaks = 10.5) +
  geom_segment(data = data_frame(q=123,p=10.5),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=127,p=10.5),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2)+
  labs(title="Perfectly Elastic Supply (\U221E)", x="Q", y="P")+
  geom_text(aes(x = 130, y = 10.7, label = "S[2]"), parse = TRUE, size =8, color = '#e0e0e0')
grafik12
```
an illustration of **perfectly elastic supply**
:::
::::

## Two extreme cases of elasticity

- In a **perfectly inelastic supply** case, the market can't supply more even if the price increase. Toll road is like this: Building more toll road is expensive and takes time. Most times this is monopolized by the government.

- In a **perfectly elastic supply**, the market supply exactly zero when the price go up just by a little. A highly competitive market where all firms operate at the margin may behave this way.

- Again, most times, the supply curve is somewhere in between.

## What changes supply elasticity?

- A market tend to have a highly elastic price elasticity of supply when **inputs are abundant** and can be converted in and out of producion at a very low cost.

- Matters, as price elasticity of supply tend to be higher in the long run. Increasing production capacity may takes time since investing in a new land, building or machine requires **time**.
  - Some goods are even impossible to be increased, hence have zero elasticity. Radio spectrum is one example.
  
## European farm surplus

European Union **subsidised** their farmers heavily. They probably knew that subsidies create huge **surplus** (remember last week), but they thought the surplus would not be as big since farming land is **limited**.

However, EU farmers were able to expand production using things like fertilizers and pesticides, which are **readily available inputs**.

# The benefits and costs of taxation

## Excise Tax (_cukai_)

- The government often tax goods in the form of an excise tax.

- In Indonesia, some goods have an excise to its purchase:
  - Cigarettes & alcohol
  - Luxury goods (PPnBM)

- You can say that a value-added tax is another form of an excise tax.

## How tax works

- Tax acts like a price ceiling plus plus floor:
  - The government set a higher price for consumers to pay
  - At the same time, the producers are forced to produced at a lower price.
  
- A reminder: in a **price floor** scenario, surplus from consumer transferred to the producer, while in a **price ceiling**, surplus transferred from producers to consumers.

- in tax case, CS and PS are transferred to the government.

## The impact of tax {.m}

```{r grafiks3, echo=FALSE, message=FALSE, warning=FALSE}
supply <- function(q) (q/12.75)
supply1 <- function(q) (q/41.3)^(1/0.5)
supply2 <- function(q) (q/2.45)^(1/1.7)

x<-120:150
eqq1 <- uniroot(function(x) demand1(x) - supply(x), range(x))$root
eqq2 <- uniroot(function(x) demand2(x) - supply(x), range(x))$root
eqq3 <- uniroot(function(x) demand(x) - supply1(x), range(x))$root
eqq4 <- uniroot(function(x) demand(x) - supply2(x), range(x))$root
eqq5 <- uniroot(function(x) demand1(x) - supply1(x), range(x))$root

# Equilibrium price
eqp1 <- supply(eqq1)
eqp2 <- supply(eqq2)
eqp3 <- supply(eqq3)
eqp4 <- supply(eqq4)
eqp5 <- supply(eqq5)

grafiks5 <- graph +
  stat_function(fun = supply1,size=1.1, color='red') +
  stat_function(fun = demand1,size=1.1, color='cyan') +
  scale_x_continuous(limits= c(120, 150), 
                     breaks = c(130,eqq5), labels = c(expression(Q[T]),expression(Q[E]))) +
  scale_y_continuous(limits = c(9,12),
                     breaks = c(supply1(130),eqp5,demand1(130)), labels = c(expression(P[P]),expression(P[E]),expression(P[C]))) +
  geom_point(aes(x=eqq5,y=eqp5),color='white',size=4)+
  geom_point(aes(x=130,y=supply1(130)),color='white',size=4)+
  geom_point(aes(x=130,y=demand1(130)),color='white',size=4)+
  geom_segment(data = data_frame(q=130,p=supply1(130)),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130,p=supply1(130)),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130,p=demand1(130)),
               aes(x = q, y = supply1(130), xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130,p=demand1(130)),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2)+
  geom_segment(data = data_frame(q=eqq5,p=eqp5),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2)+
  geom_segment(data = data_frame(q=eqq5,p=eqp5),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2)+
  geom_ribbon(aes(x=seq(0,130,1), ymin=supply1(130), ymax=eqp5, fill="C"),alpha = 0.4,show.legend=FALSE)+
  geom_ribbon(aes(x=seq(0,130,1), ymin=eqp5, ymax=demand1(130), fill="A"),alpha = 0.4,show.legend=FALSE)+
  geom_ribbon(aes(x=seq(130,eqq5,1), ymin=supply1(seq(130,eqq5,1)), ymax=eqp5, fill="B"),alpha = 0.4,show.legend=FALSE)+
  geom_ribbon(aes(x=seq(130,eqq5,1), ymin=eqp5, ymax=demand1(seq(130,eqq5,1)), fill="D"),alpha = 0.4,show.legend=FALSE)+
  labs(x="Q (1000 cars/day)", y="P (1000 IDR)")+
  geom_text(aes(x = 145, y = 9.3, label = "D"), parse = TRUE, size =8, color = '#e0e0e0')+
  geom_text(aes(x = 145, y = 11.7, label = "S"), parse = TRUE, size =8, color = '#e0e0e0')+
  annotate("text", x = 125, y = 10.2, label = "C", size=9)+
  annotate("text", x = 131.5, y = 10.3, label = "D", size=9)+
  annotate("text", x = 125, y = 10.7, label = "A", size=9)+
  annotate("text", x = 131.5, y = 10.7, label = "B", size=9)

grafiks5

```

## The impact of taxation {.m}

:::: columns

::: {.column width=47%}
```{r grafiks6, echo=FALSE, message=FALSE, warning=FALSE}
#| fig-width: 7
#| fig-height: 7
grafiks5
```
:::
::: {.column width=47%}
- The market price is $P=P_E$

- The price paid by consumers is $P=P_C$

- Producer receive payment at $P=P_P$

- The tax per purchased good paid to the government is $t=P_C-P_P$
:::
::::

---

## The impact of taxation {.m}

:::: columns
::: {.column width=47%}
```{r grafiks7, echo=FALSE, message=FALSE, warning=FALSE}
#| fig-width: 7
#| fig-height: 7
grafiks5
```
:::

::: {.column width=47%}
- Consumer lose $A+B$

- Producer lose $C+D$

- The government collects tax revenue $TR=T \times Q_T$

- This is equals to $TR=A+C$

- $B+D=DWL$
:::
::::

## The impact of taxation

- As we learn, the tax revenue and the size of DWL depends highly in demand elasticity and supply elasticity of the good.

- The less elastic the demand and supply of a good, the higher revenue gained from taxing the good.
  - i.e., price effect > quantity effect.
  
- DWL also lower when the demand and/or supply of the good is highly inelastic.

---

:::: columns
::: {.column width=47%}
```{r grafiks1, echo=FALSE, message=FALSE, warning=FALSE}

grafiks1 <- graph +
  stat_function(fun = supply,size=1.1, color='red') +
  stat_function(fun = demand1,size=1.1, color='cyan') +
  scale_x_continuous(limits= c(120, 150), 
                     breaks = c(130,eqq1), labels = c(expression(Q[T]),expression(Q[E]))) +
  scale_y_continuous(limits = c(9,12),
                     breaks = c(supply(130),eqp1,demand1(130)), labels = c(expression(P[P]),expression(P[E]),expression(P[C]))) +
  geom_point(aes(x=eqq1,y=eqp1),color='white',size=4)+
  geom_point(aes(x=130,y=supply(130)),color='white',size=4)+
  geom_point(aes(x=130,y=demand1(130)),color='white',size=4)+
  geom_segment(data = data_frame(q=130,p=supply(130)),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130,p=supply(130)),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130,p=demand1(130)),
               aes(x = q, y = supply(130), xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130,p=demand1(130)),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2)+
  geom_segment(data = data_frame(q=eqq1,p=eqp1),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2)+
  geom_segment(data = data_frame(q=eqq1,p=eqp1),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2)+
  geom_ribbon(aes(x=seq(0,130,1), ymin=supply(130), ymax=demand1(130), fill="TR"),alpha = 0.4,show.legend=FALSE)+
  geom_ribbon(aes(x=seq(130,eqq1,1), ymin=supply(seq(130,eqq1,1)), ymax=demand1(seq(130,eqq1,1)), fill="DWL"),alpha = 0.4,show.legend=FALSE)+
  annotate("text", x = 125, y = 10.6, label = "Revenue", size=8)+
  annotate("text", x = 125, y = 10.8, label = "Tax", size=8)+
  annotate("text", x = 137, y = 11.6, label = "DWL", size=8)+
  annotate("segment", x = 131.2, xend = 135.2, y = 10.7, yend = 11.3,
           arrow = arrow(length = unit(2, "lines")), size=1, colour = "gray")+
  labs(title="(a) Inelastic demand (<1)", x="Q (1000 cars/day)", y="P (1000 IDR)")+
  geom_text(aes(x = 145, y = 9.3, label = "D"), parse = TRUE, size =8, color = '#e0e0e0')+
  geom_text(aes(x = 145, y = 11.1, label = "S"), parse = TRUE, size =8, color = '#e0e0e0')

###################################################################

grafiks2 <- graph +
  stat_function(fun = supply,size=1.1, color='red') +
  stat_function(fun = demand2,size=1.1, color='cyan') +
  scale_x_continuous(limits= c(120, 150), 
                     breaks = c(130,eqq2), labels = c(expression(Q[T]),expression(Q[E]))) +
  scale_y_continuous(limits = c(9,12),
                     breaks = c(supply(130),eqp2,demand2(130)), labels = c(expression(P[P]),expression(P[E]),expression(P[C]))) +
  geom_point(aes(x=eqq2,y=eqp2),color='white',size=4)+
  geom_point(aes(x=130,y=supply(130)),color='white',size=4)+
  geom_point(aes(x=130,y=demand2(130)),color='white',size=4)+
  geom_segment(data = data_frame(q=130,p=supply(130)),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130,p=supply(130)),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130,p=demand2(130)),
               aes(x = q, y = supply(130), xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130,p=demand2(130)),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2)+
  geom_segment(data = data_frame(q=eqq2,p=eqp2),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2)+
  geom_segment(data = data_frame(q=eqq2,p=eqp2),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2)+
  geom_ribbon(aes(x=seq(0,130,1), ymin=supply(130), ymax=demand2(130), fill="TR"),alpha = 0.4,show.legend=FALSE)+
  geom_ribbon(aes(x=seq(130,eqq2,1), ymin=supply(seq(130,eqq2,1)), ymax=demand2(seq(130,eqq2,1)), fill="DWL"),alpha = 0.4,show.legend=FALSE)+
  labs(title="(b) elastic demand (>1)", x="Q (1000 cars/day)", y="P (1000 IDR)")+
  geom_text(aes(x = 145, y = 10.3, label = "D"), parse = TRUE, size =8, color = '#e0e0e0')+
  geom_text(aes(x = 145, y = 11.7, label = "S"), parse = TRUE, size =8, color = '#e0e0e0')

########################################################################

grafiks3 <- graph +
  stat_function(fun = supply1,size=1.1, color='red') +
  stat_function(fun = demand,size=1.1, color='cyan') +
  scale_x_continuous(limits= c(120, 150), 
                     breaks = c(130,eqq3), labels = c(expression(Q[T]),expression(Q[E]))) +
  scale_y_continuous(limits = c(9,12),
                     breaks = c(supply1(130),eqp3,demand(130)), labels = c(expression(P[P]),expression(P[E]),expression(P[C]))) +
  geom_point(aes(x=eqq3,y=eqp3),color='white',size=4)+
  geom_point(aes(x=130,y=supply1(130)),color='white',size=4)+
  geom_point(aes(x=130,y=demand(130)),color='white',size=4)+
  geom_segment(data = data_frame(q=130,p=supply1(130)),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130,p=supply1(130)),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130,p=demand(130)),
               aes(x = q, y = supply1(130), xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130,p=demand(130)),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2)+
  geom_segment(data = data_frame(q=eqq3,p=eqp3),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2)+
  geom_segment(data = data_frame(q=eqq3,p=eqp3),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2)+
  geom_ribbon(aes(x=seq(0,130,1), ymin=supply1(130), ymax=demand(130), fill="TR"),alpha = 0.4,show.legend=FALSE)+
  geom_ribbon(aes(x=seq(130,eqq3,1), ymin=supply1(seq(130,eqq3,1)), ymax=demand(seq(130,eqq3,1)), fill="DWL"),alpha = 0.4,show.legend=FALSE)+
  labs(title="(c) inelastic supply (<1)", x="Q (1000 cars/day)", y="P (1000 IDR)")+
  geom_text(aes(x = 145, y = 9.3, label = "D"), parse = TRUE, size =8, color = '#e0e0e0')+
  geom_text(aes(x = 145, y = 11.7, label = "S"), parse = TRUE, size =8, color = '#e0e0e0')

#####################################################################

grafiks4 <- graph +
  stat_function(fun = supply2,size=1.1, color='red') +
  stat_function(fun = demand,size=1.1, color='cyan') +
  scale_x_continuous(limits= c(120, 150), 
                     breaks = c(130,eqq4), labels = c(expression(Q[T]),expression(Q[E]))) +
  scale_y_continuous(limits = c(9,12),
                     breaks = c(supply2(130),eqp4,demand(130)), labels = c(expression(P[P]),expression(P[E]),expression(P[C]))) +
  geom_point(aes(x=eqq4,y=eqp4),color='white',size=4)+
  geom_point(aes(x=130,y=supply2(130)),color='white',size=4)+
  geom_point(aes(x=130,y=demand(130)),color='white',size=4)+
  geom_segment(data = data_frame(q=130,p=supply2(130)),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130,p=supply2(130)),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130,p=demand(130)),
               aes(x = q, y = supply2(130), xend = q, yend = p), lty = "dotted", size=2) +
  geom_segment(data = data_frame(q=130,p=demand(130)),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2)+
  geom_segment(data = data_frame(q=eqq4,p=eqp4),
               aes(x = 120, y = p, xend = q, yend = p), lty = "dotted", size=2)+
  geom_segment(data = data_frame(q=eqq4,p=eqp4),
               aes(x = q, y = 9, xend = q, yend = p), lty = "dotted", size=2)+
  geom_ribbon(aes(x=seq(0,130,1), ymin=supply2(130), ymax=demand(130), fill="TR"),alpha = 0.4,show.legend=FALSE)+
  geom_ribbon(aes(x=seq(130,eqq4,1), ymin=supply2(seq(130,eqq4,1)), ymax=demand(seq(130,eqq4,1)), fill="DWL"),alpha = 0.4,show.legend=FALSE)+
  labs(title="(d) elastic supply (>1)", x="Q (1000 cars/day)", y="P (1000 IDR)")+
  geom_text(aes(x = 145, y = 9.3, label = "D"), parse = TRUE, size =8, color = '#e0e0e0')+
  geom_text(aes(x = 145, y = 11.3, label = "S"), parse = TRUE, size =8, color = '#e0e0e0')

grafiks1
grafiks3

```
:::

::: {.column width=47%}
```{r grafiks2, echo=FALSE, message=FALSE, warning=FALSE}
grafiks2
grafiks4

```
:::
::::

## Implication

- If the government wants to collect revenue, its best to impose a tax on a good with low elasticities.

- If the government wants to change behaviour, tax a good with high elasticities.

- Taxing cigarettes to discourage smoking maybe ineffective
  - however, it can be used to gain revenue.
  
- Highly inelastic goods may suggest it's a necessity: people may be unhappy with the tax.

## Implication

- Same goes with the business people: be careful in increasing price if your product is highly elastically demanded.

- Next week, we will have a look a bit more detailed on the supply curve: 
  - what drives the elasticity of supply.
  - Different types of cost.
  
## Some recap before da quiz

- Elasticities is very important in determining surpluses: CS, PS, revenues.

- Elasticities can be estimated (albeit hard).

- Elasticities vary: $0 \leq 1 \leq \infty$ .

- Many factors influence elasticities.

- You should be able to use elasticity formula and can draw how tax is calculated and drawn.

## Thank You

<iframe src="https://giphy.com/embed/NIsWtysJUrvTG" width="480" height="320" frameBorder="0" class="giphy-embed" allowFullScreen></iframe><p><a href="https://giphy.com/gifs/cute-NIsWtysJUrvTG">via GIPHY</a></p>