---
title: "The New Trade Model"
subtitle: "ECES905205 Meeting 5"
date: "September 2022"
author: "I Made Krisna Gupta"
format: 
  revealjs:
    theme: [simple,../mytheme.scss]
    chalkboard: true
    incremental: false
    transition: convex
    background-transition: fade
    slide-number: true
  beamer: default
---

## Latest iteration

- Gains from trade so far:

    - different technology.
    - different factor of endowment.
    
- Today:

    - product variety.
    - productivity improvement from selection.

## Internal economies of scale {.m}

- We relax perfect market a bit: non-trivial fixed cost.

- Consequently, average cost goes down with scale.

- Meaning, a market can only host a limited number of firms.

- In reality, fixed cost come in many firms: R&D, network, machines, etc.

## Internal economies of scale

- In reality, there are some industries where the global market only shared by small number of MNCs.

    - semiconductor, aeroplane, software, soft drink.
    
- In this context, trade brings more gain:

    - the most efficient firms will dominate market and push cost down.
    
    - inefficient firms exit.

## example: mobile phone

<div id="mobile_vendor-ww-yearly-2021-2021" width="600" height="400" style="width:600px; height: 400px;"></div><!-- You may change the values of width and height above to resize the chart --><p>Source: <a href="https://gs.statcounter.com/vendor-market-share/mobile/worldwide/#yearly-2021-2021-bar">StatCounter Global Stats - Device Vendor Market Share</a></p><script type="text/javascript" src="https://www.statcounter.com/js/fusioncharts.js"></script><script type="text/javascript" src="https://gs.statcounter.com/chart.php?mobile_vendor-ww-yearly-2021-2021&chartWidth=600"></script>

## Monopolistic competition

- In monopolistic competition, firms have some ability to control price.

- Condition which enable this:

    - Few number of players.
    
    - DIfferentiated products.
    
- Product differentiation may come from various aspect:

    - Technology: smartphones, OS, recipe.
    
    - Network effect: social media, apps store, Gojek/Grab.

## Monopoly: a brief review

- Assume an industry face demand $Q=a-bP$
- $P$ is not exogen (why?), thus rearrange:
  - $P=\frac{a}{b}-\frac{Q}{b}$
- Revenue in the industry $R=Q\times P$, subs $P$:
  - $R=Q \times \left(\frac{a}{b}-\frac{Q}{b}\right)$
  - $MR=\frac{a}{b}-2\frac{Q}{b}=P-\frac{Q}{b}$

## Monopoly: a brief review

- Total cost: $C=F+cQ$
- Marginal cost: $c$
- Average cost: $\frac{F}{Q}+c$
- as $Q \rightarrow \infty$, $\frac{F}{Q} \rightarrow 0$
- The larger F, the larger Q is required.

---

## The cost curve

AC2 has a higher fixed cost, thus requires more Q to reach "flat-ish" part of the cost curve.

```{r}
#| output: false
library(tidyverse)

plot<-ggplot()+theme_classic()+
  theme(panel.background = element_rect(fill = '#f0f1eb'),
        plot.background = element_rect(fill = '#f0f1eb'),
        axis.title.x = element_text(hjust=1),
        axis.title.y = element_text(angle=0,hjust=1,vjust=1),
        panel.grid=element_blank(),
#        axis.text.x=element_blank(),
#        axis.text.y=element_blank(),
#        axis.ticks.x=element_blank(),
#        axis.ticks.y=element_blank(),
        legend.position=c(1,1),
        legend.justification = c(1,1),
        legend.spacing=unit(0,"cm"),
        legend.margin=margin(0,0,0,0))
```

```{r}
#| label: fig1
#| warning: false
AC1 <- function(q) 1.5/q+1
AC2 <- function(q) 10/q+1
plot+
  stat_function(fun = AC1,size=1.1, color='red') +
  stat_function(fun = AC2,size=1.1, color='blue') +
  geom_hline(yintercept=1,size=1.1,color='black')+
  labs(x="Q", y="C",color=NULL,fill=NULL)+
  annotate("text",x=10,y=1.3,label="AC1")+
  annotate("text",x=20,y=2,label="AC2")+
  annotate("text",x=10,y=0.8,label="MC")+
  scale_x_continuous(expand=c(0,0),limits = c(0,30))+
  scale_y_continuous(expand=c(0,0),limits = c(0,5))+
  theme(      axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.x=element_blank(),
        axis.ticks.y=element_blank())
```

## Monopoly Q

::: columns
::: {.column width="47%"}

```{r}
#| label: fig2
#| warning: false
#| fig-width: 5
#| fig-height: 4

demand<- function(q) 5.5-q
mr<- function(q) 5.5-q-q
z<-seq(0, 4.5/2, 0.01)
AC1 <- function(q) 2/q+1

plot+
  stat_function(fun = AC1,size=1.1, color='red') +
  stat_function(fun = demand,size=1.1, color='blue') +
  stat_function(fun = mr,size=1.1, color='#008080') +
  geom_hline(yintercept=1,size=1.1,color='black')+
  labs(x="Q", y="C,P",color=NULL,fill=NULL)+
  geom_ribbon(aes(x = z, ymin = AC1(4.5/2), ymax=demand(4.5/2)), alpha = 0.25) +
  annotate("point",x=4.5/2,y=1)+
  annotate("segment",x=4.5/2,y=0,xend=4.5/2,yend=demand(4.5/2),lty="dotted",size=1.1)+
  annotate("segment",x=0,y=demand(4.5/2),xend=4.5/2,yend=demand(4.5/2),lty="dotted",size=1.1)+
  annotate("segment",x=0,y=AC1(4.5/2),xend=4.5/2,yend=AC1(4.5/2),lty="dotted",size=1.1)+
  annotate("text",x=4.5,y=1.8,label="AC")+
  annotate("text",x=4.9,y=1.1,label="MC")+
  annotate("text",x=4.8,y=0.5,label="Demand")+
  annotate("text",x=2.7,y=0.5,label="MR")+
  scale_x_continuous(expand=c(0,0),limits = c(0,5),breaks = 4.5/2,labels="Qm")+
  scale_y_continuous(expand=c(0,0),limits = c(0,5),breaks=c(AC1(2.25),demand(2.25)),
                     labels=c("ACm","Pm"))
  
```
:::
::: {.column width="47%"}
- monopolist akan memaksimalkan profit di MR=MC
- pastinya akan ngecharge P>AC
- profit $=(P-AC)Q$
:::
:::

## Monopolistic competition

- Characteristics of monopolistic competition industries:
  - Firms can differentiate.
  - Rival's price is exogenous.
  - free entry.

- We also introduce high fixed cost!

    - while entry is free, a firm must pay some fixed cost.
    
- Can you mention some examples?

## Karakteristik monopolistic

$$
Q=S\times\left[1/n-b(P-\bar{P})\right]
$$
- Q = total sales of individual firm
- S = total sales of the industry
- n = number of firms
- b = elasticity of supply
- P = firms' price
- $\bar{P}$ = rival's price index

## Karakteristik monopolistic 

$$
Q=S\times\left[1/n-b(P-\bar{P})\right]
$$

- A firm's sales go up if total sales of the industry goes up and/or rivals' price go up.
- A firm's sales goes down if the number of firms go up and its own price go up.
- All firms are identic: $P=\bar{P}$, all firms share the market equally $S/n$
- $AC=\frac{F}{Q}+c \Rightarrow AC=n\frac{F}{S}+c$

## Monopolistic characteristics

$AC=n\frac{F}{S}+c$

- Average cost goes up as numbers of firm go up. This is because firms' market share goes down. berkurang.
- As the overall pie goes up, average cost go down: the fixed cost can be shared among more sales.
- With a linear demand $Q=a-bP$, firms produce at MR=MC
- $P-\frac{Q}{Sb}=c \rightarrow \frac{Q}{Sb}=P-c \rightarrow P-c=\frac{1}{nb}$

## Monopolistic Characteristics

$P-c=\frac{1}{nb}$

- As the number of firms go up, $P-c$ (markups) goes down.
  - Since c is exogenous, reduction of markups is driven by $P \Downarrow$
  - More firms, less market share, P must go up to avoid loss.
- Firms will keep coming in as long as there's profit.
- In the long-run, number of firms settles at $P=AC$

## Monopolistic

::: columns
::: {.column width="47%"}

```{r}
#| label: fig3
#| warning: false
#| fig-width: 5
#| fig-height: 4

DD<- function(q) 9/q+1
CC<- function(q) q+1

plot+
  stat_function(fun = DD,size=1.1, color='blue') +
  stat_function(fun = CC,size=1.1, color='red') +
  labs(x="n", y="C,P",color=NULL,fill=NULL)+
  scale_x_continuous(expand=c(0,0),limits = c(0,10),breaks = 3,labels="n*")+
  scale_y_continuous(expand=c(0,0),limits = c(0,10),breaks=4,labels="P*,AC*")+
  annotate("text",x=7,y=7,label="AC(n,.)")+
  annotate("text",x=7,y=3,label="P(n,.)")
  
```
:::
::: {.column width="47%"}
- Any higher than $n^*$, firms cost is too much.
- Any lower than $n^*$, mark-up will go up, incentives for new entrant to join.
:::
:::

## Trade & monopolistic

::: columns
::: {.column width="47%"}

```{r}
#| label: fig4
#| warning: false
#| fig-width: 5
#| fig-height: 4

DD<- function(q) 9/q+1
DDD<- function(q) 18/q+1
CC<- function(q) q+1
CCC<- function(q) q/2+1

plot+
  stat_function(fun = DD,size=1.1, color='blue') +
  stat_function(fun = CC,size=1.1, color='red') +
  stat_function(fun = CCC,size=1.1, color='red',lty='dashed') +
  labs(x="n", y="C,P",color=NULL,fill=NULL)+
  scale_x_continuous(expand=c(0,0),limits = c(0,10),breaks =c(3,sqrt(18)),labels=c("n1","n2"))+
  scale_y_continuous(expand=c(0,0),limits = c(0,10),breaks=c(4,DD(sqrt(18))),labels=c("P1,AC1","P2,AC2"))+
  annotate("text",x=7,y=7,label="AC(n,.)")+
  annotate("text",x=7,y=3,label="P(n,.)")

  
```

$$
AC=n\frac{F}{S}+c \\
P=\frac{1}{nb}+c
$$

:::
::: {.column width="47%"}
- Opening to trade increases the market size.
  - $S \uparrow$ increase AC (less share per firm), but pricing is unaffected.
  - New equilibrium is lower price and more firms in the market.
- More firms->more varieties.

:::
:::

## Nummeric examples

- Let an automotive industries as follows:

$$
Q=S\times[(1/n)-(1/30000)\times(P-\bar{P})] \\
$$

- with this cost structures:

$$
AC=750000000/Q+5000
$$

- Let home sells 900k vehicles a year while foreign produces 1,600k.

## Numeric examples

```{r}
#| label: fig5
#| warning: false

library("patchwork")
DA<- function(q) 5000+1/(1/30000*q)
CA<- function(q) 750000000*q/900000+5000
CB<- function(q) 750000000*q/1600000+5000
C3<- function(q) 750000000*q/2500000+5000


a<-plot+
  stat_function(fun = DA,size=1.1, color='blue') +
  stat_function(fun = CA,size=1.1, color='red') +
  labs(x="n", y="C,P",color=NULL,fill=NULL)+
  scale_x_continuous(expand=c(0,0),limits = c(0,12))+
  scale_y_continuous(expand=c(0,0),limits = c(0,36000))+
  ggtitle("Home")

b<-plot+
  stat_function(fun = DA,size=1.1, color='blue') +
  stat_function(fun = CB,size=1.1, color='red') +
  labs(x="n", y="C,P",color=NULL,fill=NULL)+
  scale_x_continuous(expand=c(0,0),limits = c(0,12))+
  scale_y_continuous(expand=c(0,0),limits = c(0,36000))+
  ggtitle("Foreign")

c<-plot+
  stat_function(fun = DA,size=1.1, color='blue') +
  stat_function(fun = C3,size=1.1, color='red') +
  labs(x="n", y="C,P",color=NULL,fill=NULL)+
  scale_x_continuous(expand=c(0,0),limits = c(0,12))+
  scale_y_continuous(expand=c(0,0),limits = c(0,36000))+
  ggtitle("Trade")

a+b+c
```

## Contoh numerik

|     | Home, no trade | Foreign, no trade | Integrated market |
| --- | -------------- | ----------------- | ----------------- |
| Output | 900,000 | 1,600,000 | 2,500,000 |
| n | 6 | 8 | 10 |
| Output per firm | 150,000 | 200,000 | 250,000 |
| AC=P | 10,000 | 8,750 | 8,000 |

## Highlighted feature

- Even when both countries have the same technology (identical cost structure), trade brings positive impact.
- Monopolistic competition is benefited from **intra-industry trade**.
  - Export car but also import car.
  - Think of Toyota and Ford.
- more car variety: from 6 & 8 brands to 10 brands for both countries.

## Highlighted feature

- Note that intra-industry trade is more beneficial for Home
  - price reduces by 2k, foreign only by 750.
- This is because home country have a smaller market.
  - The smaller the market, the less fixed cost is shared.
  - This is why small countries like Singapore, Malaysia, Vietnam and Thailand pursue FTAs vigorously.

## Firms response to trade

- From 14 brands of car, only 10 survives.
- These 10 brands get more market share, hence able to reduce cost.
- Since all firms are identical in our analysis, we do not care who out and who're staying.
- However, in the real world, firms are not identical: 4 firms who exit must be the worst performers.


## Firms response to trade

- Market integration forced firms with worse cost structure to exit.
- If the fixed cost to enter the industry is the same, cost variation will come from variable cost.
- Then there's a threshold $c^*$ where all firms with $c>c^*$ exit.

## More gain from trade

- This, in turn, adds to the gain from trade:

  - worst firms will exit, only firms with $c<c^*$ stays.
  - meaning, price index further even more from lower average variable cost.
  - If 5 worst students drop out, average GPA will go up.
  
- Firms who exit will be absorbed by the stayers: either naturally or through acquisition/joint-venture.

## Export cost

- We can associate a fixed cost with export cost:
  - establish presence, market research, finding partners.
  - Only the most productive firms join the global market.
- Opposite argument is also true:
  - Learning by doing: As the firm keeps on exporting, they learn from the global market and become much more productive.
  
## Multinationals and FDI

- In general, FDI is considered more costly than ekspor.
- Greenfield FDI: Brand new investment.
- Brownfield FDI: purchasing someone else's asset.
- Horizontal FDI: Replicating domestic success.
- Vertical FDI: global value chain.

## Multinationals and FDI

- Horizontal FDI is done under high trade cost.
  - the goal is market access. Usually will not export to other countries.
- Vertical FDI: the goal is to supress cost.
  - Low-tech countries can export high-tech goods.
  - Requires low trade cost, and possible large intermediate imports.
  - leads to intrafirm trade.
  
## Intrafirm Trade {.m}

- Analyzing between-country trade is getting less relevant as the importance of intrafirm trade improves.
- Half of total imports of US are intrafirm. However, around 10% of US trade with Indonesia is intrafirm.
- The difference between exporters and non-exporters is mostly driven by firms that both import and export. This is true also in Indonesia.
- Amiti and Konings (2007) show that a reduction of import tariff by half improves productivity of an importing firm by 12%, which is larger than the reduction of output tariff by destination countries.
- Channels: foreign technology embeded in the inputs, variety of inputs, cheaper and higher quality inputs.
- Others try to find more evidence for Indonesia (e,g., Pane and Patunru 2019, Gupta 2020), but data availability in firm level remains a huge challenge.

## {background-image="wew.jpg"}