---
title: "Indonesia in BRICS: a discussion"
subtitle: ""
date: "19 February 2025"
date-format: "long"
author: "Krisna Gupta"
format: 
  revealjs:
    theme: [simple,mytheme.scss]
    chalkboard: true
    incremental: false
    transition: convex
    background-transition: fade
    slide-number: true
  beamer:
    latex-tinytex: true
    latex-auto-install: true
  pptx: default
---

```{r}
library(tidyverse)
library(readxl)
library(WDI)
library(scales)

graph<-function(dataset, ...) {
  dataset |>
  ggplot(aes(...))+
  theme_classic() +
  theme(panel.background = element_rect(fill = "#f0f1eb",
                                        colour = "#f0f1eb"),
        plot.margin = margin(0,0,0,0),
        plot.background = element_rect(fill = "#f0f1eb"),
        legend.key = element_rect(fill = "#f0f1eb",linetype = "blank"),
        legend.background = element_rect(fill="#f0f1eb"),
        axis.text=element_text(size=16))
}
```

## Intro

- Indonesia surprises many when it suddenly joined BRICS.

    - something avoided by the previous administration.

- Main reasons: alternatives to the western-led global order.

    - Does Indonesia really needs to join BRICS?

- BRICS is a part of a bigger conversation: the end(?) of a unipolarity world.

    - Move aside, economists!

## Why?

- Export market access.

- Fairer global order, more agency.

- Alternative financing for development & agency.

- Alternatives to the dollar.

- Alternative source of technology: climate, food, defense.

## Why not?

- BRICS is extremely diverse.

- Some of the previous arguments may not be true.

- Many of the previous arguments may be achievable without joining BRICS.

- BRICS may not matter in reality, but it may matter in Trump's mind.

- Out of the west, into the dragon?

## GDP {.m}

```{r}
indi<-c(            # membuat dictionary
  "PDB"="NY.GDP.MKTP.CD" # NY.GDP.MKTP.PP.CD vs NY.GDP.MKTP.CD
)

dat<-WDI(           # Menarik data World Bank
  country=c("BRA","RUS","IND","CHN","ZAF","EGY","ETH","IRN","ARE","IDN","WLD"),
  indicator=indi,
  start=2023,end=2023
)

wld <- dat |> filter(iso3c == "WLD") %>% select(PDB) %>% pull()

dat<-dat|>filter(iso3c!="WLD")
brics<-sum(dat$PDB)
bricsf<-brics/wld
dat$frac1<-dat$PDB/brics*100
dat<-dat|>arrange(frac1)|>mutate(country=factor(country,country))

graph(dat,x=country,y=frac1)+geom_bar(stat="identity")+coord_flip()+
  labs(x="",y="%",caption="source: WDI",
       title="Share of GDP (PPP) within BRICS membership")
```

In 2023, BRICS member represents around 38% of the world's GDP PPP. G7 is 27%. It is, however, unorthodox, in the sense it doesn't have a secretariat. China's size in BRICS > US' in the G7.

## Export (left) & import, 2023

:::: columns
::: {.column width=49%}
![](ekspor.svg)
:::
::: {.column width=49%}
![](impor.svg)
:::
::::

## Trade Balance {.m}

:::: columns
::: {.column width=49%}

```{r}
#| fig-width: 5
#| fig-length: 5
indi<-c(            # membuat dictionary
  "ca"="BN.GSR.GNFS.CD" # BN.GSR.GNFS.CD vs BN.CAB.XOKA.CD
)

CAA<-WDI(           # Menarik data World Bank
  country=c("BRA","RUS","IND","CHN","ZAF","EGY","ETH","IRN","ARE","IDN"),
  indicator=indi
)

bricsca<-CAA|>drop_na()
swt<-bricsca|>group_by(year)|>summarize(total=sum(ca))

swt$total<-swt$total/1e9

graph(swt,x=year,y=total)+geom_line(linewidth=1.1)+
  scale_y_continuous(labels = label_comma(accuracy=1))+
  geom_hline(yintercept=0)+
  labs(title="Total net export of goods and services of\nBRICS current members",
       x="",y="Billion current USD",caption="source:IMF via WDI")+
  theme(legend.position="bottom")
```

On aggregate, BRICS's export is larger than its import. Collectively, they all need other markets.

:::
::: {.column width=49%}

```{r}
#| fig-width: 5
#| fig-length: 5
indi<-c(            # membuat dictionary
  "ca"="BN.GSR.GNFS.CD" # BN.GSR.GNFS.CD vs BN.CAB.XOKA.CD
)
CA<-WDI(           # Menarik data World Bank
  country=c("BRA","RUS","IND","CHN","ZAF","IDN"),
  indicator=indi
)

CA$ca<-CA$ca/1e9

graph(CA,x=year,y=ca,color=country)+geom_line(linewidth=1.1)+
  scale_y_continuous(labels = label_comma(accuracy=1))+
  geom_hline(yintercept=0)+
  labs(title="Net export of the original BRICS+Indonesia",
       x="",y="Current billion USD",caption="source:IMF via WDI")+
  theme(legend.position="bottom")
```

India is the largest deficit countries, but it won't act as 'the buyer' in BRICS.

:::
::::

## BRICS-USA relationship {.m}

```{r}
indi<-c(            # membuat dictionary
  "ca"="BN.GSR.GNFS.CD" # BN.GSR.GNFS.CD vs BN.CAB.XOKA.CD
)
CA<-WDI(           # Menarik data World Bank
  country=c("BRA","RUS","IND","CHN","ZAF","IDN","USA"),
  indicator=indi
)

CA$ca<-CA$ca/1e9

graph(CA,x=year,y=ca,color=country)+geom_line(linewidth=1.1)+
  scale_y_continuous(labels = label_comma(accuracy=1))+
  geom_hline(yintercept=0)+
  labs(x="",y="Current billion USD",caption="source:IMF via WDI")
```

Decoupling won't happen without radical changes in relative prices and productivity shock, a lose-to-all situation (economically)

## Dedollarization

```{r}
cofer<-read_csv("COFER.csv")|>
  select(!`Country Code`)|>select(!`Indicator Code`)|>
  select(!`Attribute`)|>
  filter(`Indicator Name`!="Shares of Allocated Reserves, Percent")|>
  filter(`Indicator Name`!="Shares of Unallocated Reserves, Percent")|>
  filter(grepl('Shares', `Indicator Name`))|>
  filter(grepl('euro|yen|pounds sterling|U.S.|renminbi|other', `Indicator Name`))

lab<-c("Shares of Allocated Reserves, Shares of Japanese yen, Percent" = "JPY", 
               "Shares of Allocated Reserves, Shares of euro, Percent" = "EUR", 
               "Shares of Allocated Reserves, Shares of Chinese renminbi, Percent" = "CNY",
               "Shares of Allocated Reserves, Shares of pounds sterling, Percent" = "GBP",
               "Shares of Allocated Reserves, Shares of U.S. dollars, Percent" = "USD",
       "Shares of Allocated Reserves, Shares of other currencies, Percent" = "other")

wld<-cofer|>filter(`Country Name`=="World")|>
  select(!'Country Name')|>
  pivot_longer(!`Indicator Name`,values_to = "value",names_to = "year")
wld$year<-as.numeric(wld$year)
adv<-cofer|>filter(`Country Name`=="Advanced Economies")|>
  select(!'Country Name')|>
  pivot_longer(!`Indicator Name`,values_to = "value",names_to = "year")
adv$year<-as.numeric(adv$year)
emg<-cofer|>filter(`Country Name`=="Emerging and Developing Economies")|>
  select(!'Country Name')|>
  pivot_longer(!`Indicator Name`,values_to = "value",names_to = "year")
emg$year<-as.numeric(emg$year)

graph(wld,x=year,y=value,color=`Indicator Name`)+
  geom_line(linewidth=1.1)+
  scale_color_discrete(labels = lab)+
  labs(title="share of currency in the global reserves",
       y="%",x="",caption="source:COFER, IMF")
#graph(adv,x=year,y=value,color=`Indicator Name`)+geom_line()+
#  scale_color_discrete(labels = lab)
#graph(emg,x=year,y=value,color=`Indicator Name`)+geom_line()+
#  scale_color_discrete(labels = lab)
```

## Dedollarization

- Having so much dollar asset risks a sanction, alternative solution is in everyone's (non-west) interest,

- The closest alternative in BRICS is CNY. but it requires PBoC to loosen its control over the renminbi, something it probably won't give up.

- Indeed, the U.S. remains accomodative to the global saving needs.

    - Even if countries trade with other currency, it'll be converted to USD to buy US assets.
    
- another pain point in "escaping the west".
    
## Foreign Aid

```{r}
indi<-c(            # membuat dictionary
  "Aid"="DT.ODA.ODAT.CD" # NY.GDP.MKTP.PP.CD vs NY.GDP.MKTP.CD
)

aid<-WDI(           # Menarik data World Bank
  country=c("BRA","RUS","IND","CHN","ZAF","IDN"),
  indicator=indi
)

aid$Aid<-aid$Aid/1e9

graph(aid,x=year,y=Aid,color=country)+geom_line(linewidth=1.1)+
  scale_y_continuous(labels = label_comma(accuracy=0.01))+
  geom_hline(yintercept=0)+
  labs(title="Net flow of foreign aid",
       x="",y="Current billion USD",caption="source:OECD via WDI")

```

## Foreign aid

- Most of BRICS members are net recipient of foreign aid but China.

- NDB's role so far is outstripped by western-led institution.

- Indeed, Chinese role in foreign aid is increasing, especially since the mid 2010s.

- Chinese loan focuses on infrastructure development, while DAC focues on MDGs (poverty, health, etc)

- tapping to non-DAC aid doesn't require one to be a member of BRICS

## Conclusion

- Larger export market, alternative financial source and dollar alternatives may not be the feasible reasons to join BRICS.

    - worse, it risks an uncessarily costly 'decouple'.

- The non-DAC loans and grants can still be accessed even without joining BRICS.

- Many observers note the need for a platform where Indonesia can have larger role, it can be questioned whether BRICS won't be heavily influenced by China.

## What's next?

- "The west" may be in a challenging situation themselves.

- Trump isn't hostile towards BRICS: he's hostile towards everyone.

- There may be a vaccuum in the global leadership as a whole.

- It's not clear how the next global order would look like.

    - trade is not guided by comparative advantage anymore, but by politics and security.
    
## What's next

- As a patriotic Indonesian, It's time for Indonesia to take a leadership role in all its platform, helps BRICS formalized its membership, leads ASEAN, and help guiding the world.

- As a reasonably sensible Indonesian, it seems wiser to acknowledge our not-so-tough position and keep it balanced.

    - are we currently well-equipped to take a leading role?

- At least in ASEAN, Indonesia is a large player: it should act like one.

- Let's entertain the idea of the increasing role of Japan.

## References {.s}

:::: columns
::: {.column width=49%}
Associated Press. 2025. [Indonesia is admitted to the BRICS bloc of developing nations](https://www.voanews.com/a/indonesia-is-admitted-to-the-brics-bloc-of-developing-nations/7926970.html). VOA News.


Rachman, A.A. 2025. [Indonesia bergabung dengan BRICS: peluang besar dan babak baru dalam hubungan bilateral](https://theconversation.com/indonesia-bergabung-dengan-brics-peluang-besar-dan-babak-baru-dalam-hubungan-bilateral-247344). SuarAkademia TCID.

Gopinath, G & Oleg Itskhoki. 2021. Dominant Currency Paradigm: A Reviw. NBER Working Paper Series.

Intan, G. 2025. [Luhut: Indonesia tidak takut dikucilkan Amerika usai gabung BRICS](https://www.voaindonesia.com/a/luhut-indonesia-tidak-takut-dikucilkan-amerika-usai-gabung-brics/7934001.html). VOA News.

Jailani, A.K. 2025. [Indonesia’s entry into BRICS: Reshaping the global legal order](https://www.thejakartapost.com/opinion/2025/01/09/indonesias-entry-into-brics-reshaping-the-global-legal-order.html.). The Jakarta Post.

:::
::: {.column width=49%}

Bryan Tan Jia Yang. 2025. [BRICS expansion and Japan’s latent leadership in a torn Indo-Pacific](https://eastasiaforum.org/2025/02/15/brics-expansion-and-japans-latent-leadership-in-a-torn-indo-pacific/). East Asia Forum.

Mai, L. 2025. [The latest on Southeast Asia: Indonesia joins BRICS](https://www.csis.org/blogs/latest-southeast-asia/latest-southeast-asia-indonesia-joins-brics). CSIS.

Pierre van der Eng. 2017. [Why does Indonesia seem to prefer foreign aid from China?](https://eastasiaforum.org/2017/12/02/why-does-indonesia-seem-to-prefer-foreign-aid-from-china/#:~:text=Foreign%20aid%20peaked%20at%20US$2.5%20billion%20in,exceed%20new%20aid%20commitments%20from%20these%20countries.) East Asia Forum.

Economist, The. 2023. [The BRICS bloc is riven with tensions.](https://www.economist.com/international/2023/08/17/the-brics-are-getting-together-in-south-africa?ppccampaignID=17210591673&ppcadID=&gad_source=1&gclsrc=aw.ds)

Manggala, P.U. 2025. [Indonesia’s bold step into BRICS and beyond](https://eastasiaforum.org/2025/02/04/indonesias-bold-step-into-brics-and-beyond/). East Asia Forum.

Sakti, R.E. 2025. [Indonesia resmi gabung BRICS, apa tantangannya?](https://www.kompas.id/artikel/indonesia-resmi-gabung-brics-apa-tantangannya). Kompas.

:::
::::