---
title: "Trade Policy II: Beyond Tariffs, and the WTO"
subtitle: "ECEE600017 Meeting 9"
date: "4 May 2026"
author: "I Made Krisna Gupta"
format:
  revealjs:
    theme: [simple,../mytheme.scss]
    chalkboard: true
    incremental: false
    transition: convex
    background-transition: fade
    slide-number: true
  beamer: default
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
#| cache: true
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(stringr)

# Try to use DEN house style; fall back silently if ggden is not installed.
have_ggden <- requireNamespace("ggden", quietly = TRUE)
if (have_ggden) suppressMessages(library(ggden))

den_theme <- if (have_ggden && exists("theme_den", mode = "function")) {
  theme_den()
} else {
  theme_minimal(base_size = 16) +
    theme(panel.grid.minor = element_blank(),
          plot.title.position = "plot",
          plot.caption = element_text(colour = "grey40", size = 10))
}

gta_path <- "GTA NIPO - April 2026.xlsx"
gta_raw <- read_excel(
  gta_path,
  sheet = "GTA NIPO - April 2026",
  col_types = "text"
)

gta <- gta_raw |>
  transmute(
    impl   = `Implementing Jurisdiction`,
    type   = `GTA Intervention Type`,
    assess = `Initial Assessment (Change Relative to 1 Jan 2009)`,
    date   = suppressWarnings(as.Date(as.numeric(`Implementation Date`), origin = "1899-12-30"))
  ) |>
  mutate(year = as.integer(format(date, "%Y")))
```

## Content {.m}

- Last week: tariffs as the textbook trade-policy instrument:
    - Small open economy: deadweight loss; producer gain < consumer loss.
    - Large open economy: terms-of-trade gain $\rightarrow$ optimal tariff $> 0$.
    - Prisoner's dilemma: to tariff or not to tariff?

- All non-tariff can be ad-valoremed (to an extent). But what are the tools?

## Today {.m}

Three blocks plus the data:

1. **The instruments and the logic** -- quotas, VER, LCR, standards, subsidies, anti-dumping, export restrictions; why governments reach for them.

2. **The WTO** -- why a multilateral rulebook was built, what it disciplines, and why it is in trouble.

3. **What countries actually do** -- evidence from Global Trade Alert (53,000+ measures since 2009).

# Part 1 -- The instruments and the logic

## Import quota -- mechanics {.m}

- Government caps the *quantity* that can be imported.

- Effects in a small open economy with a binding quota:
    - Domestic price rises above the world price.
    - Same consumer loss and producer gain triangles as a tariff.
    - **Difference:** the rectangle that would be tariff revenue becomes a **quota rent** -- captured by whoever holds the import licence.

- Three rent destinations:
    1. Auctioned licences $\rightarrow$ government (= tariff).
    2. Allocated licences $\rightarrow$ politically chosen importers.
    3. Foreign exporters (next slide) $\rightarrow$ worst case.

## Quota vs tariff: when do they diverge? {.m}

Tariff and quota look equivalent in a static, perfectly-competitive textbook. They diverge when:

- **Demand grows.** Quota stays fixed $\rightarrow$ implicit protection rises. Tariff rate stays fixed $\rightarrow$ protection erodes.

- **Domestic firm has market power.** Quota gives the domestic monopolist a guaranteed protected market; tariff still leaves contestability from imports.

- **Rent allocation matters politically.** Quota allocations are administratively granted -- a magnet for rent-seeking.

- **Voluntary Export Restraint (VER):** quota administered by the *exporter*. Japan-US autos, 1981. The quota rent flows abroad. Strictly worse than a tariff for the home country -- yet politically convenient. Banned by the WTO Safeguards Agreement Art. 11 (1995).

## Non-Tariff Measures (NTMs): the universe {.m}

- UNCTAD MAST classification (2019) -- **16 chapters, hundreds of categories.**
    - Technical: SPS (sanitary), TBT (technical), pre-shipment inspection.
    - Non-technical: price controls, contingent measures (AD, CVD, safeguards), licences/quotas/bans, finance, distribution restrictions, government procurement, IP, rules of origin, subsidies, export measures.

- **Stylised fact:** as average tariffs fell from ~30% (1980s) to ~10% today in developing countries, the *count* of NTMs has risen.

- Hard to compare across countries because:
    - Many are not notified.
    - Tariff equivalents vary by product, firm, route.

## Local Content Requirement (LCR / TKDN) {.m}

- Mandate: a minimum share of the value (or of specified components) must be sourced domestically.

- Indonesia uses LCRs heavily:
    - **TKDN handsets** -- 4G/5G smartphones must meet ~35% TKDN to be sold; pushed Apple to negotiate, eventually invest.
    - **EV batteries / EVs** -- staircase to 80% TKDN.
    - **Solar PV, oil & gas services, government procurement.**

- **Static cost:** forces use of more expensive / lower-quality domestic inputs $\rightarrow$ tariff-equivalent often double-digit.

- **Claimed dynamic benefit:** linkages, learning, technology transfer.

- **WTO status:** TRIMs Agreement (1995) prohibits LCRs in goods. Honoured selectively; many WTO members maintain creative variants.

## Standards: SPS and TBT {.m}

- Legitimate roles: information asymmetry (consumer cannot observe quality), externalities (pesticide residue, emissions), network compatibility.

- Protectionist temptation: a strict standard *no foreign supplier can meet today* keeps competitors out, perfectly legally.

- Evidence for Indonesia (Elisabeth & Verico, 2022, *JITLP*): TBT measures have a non-trivial *dampening* effect on Indonesia's manufacturing exports -- the same instruments work both ways.

- Trade-off:

    - Tighter standard $\rightarrow$ fewer entrants $\rightarrow$ higher minimum quality, less price competition.
    - Optimal tightness depends on the externality, not on what protects domestic firms.

## Subsidies and the industrial-policy revival {.m}

- In GTA, subsidies (financial grant + state loan + state aid + trade finance) are the **single largest distortive category** by count.

- Post-2020 surge driven by:
    - **United States** -- IRA (USD 369bn, 2022), CHIPS Act (USD 52bn, 2022).
    - **EU** -- Green Deal Industrial Plan, Net-Zero Industry Act, CBAM.
    - **China** -- continued support to chips, EVs, batteries; "new productive forces".
    - **Indonesia** -- tax holidays for smelters, KEK incentives, biodiesel mandate financed by the CPO Fund.

- WTO disciplines (SCM Agreement) cover subsidies for *goods*, but actionability is high-bar and the dispute pipeline is jammed.

## Contingent protection {.m}

The "WTO-legal" way to raise protection after the schedules are bound:

- **Anti-dumping (AD)** -- duty if foreign firm sells "below normal value" and causes injury. ~4,000 distortive measures in GTA. Procedure-heavy, easy to abuse, the most-used trade defence worldwide.

- **Anti-subsidy (CVD)** -- countervailing duty against subsidised imports. ~850 measures.

- **Safeguards** -- emergency duty/quota against an *import surge* causing serious injury. Non-discriminatory in principle, time-limited, requires compensation to affected exporters.

Indonesia uses all three; Indonesia is also a *target* of dozens of foreign AD cases (palm-oil derivatives, paper, steel).

## Export-side measures {.s}

Export taxes, bans, licensing, price benchmarks. Why does an exporter restrict its own exports?

1. **Revenue.** Easy tax base on commodities (Indonesia palm-oil export levy $\rightarrow$ BPDPKS / CPO Fund).

2. **Domestic price stabilisation.** Crisis bans on rice, cooking-oil exports (2022).

3. **Downstream development.** Indonesia nickel-ore export ban (2014, then 2020) $\rightarrow$ smelter and battery investment.

4. **Strategic / geopolitical.** Critical-minerals controls (rare earths, gallium, germanium).

- Welfare logic: a *large* supplier can shift world prices to its favour -- a terms-of-trade gain. A *small* supplier just loses revenue.

- Downstream story works only if the downstream industry is genuinely viable -- otherwise it is a transfer to a subsidised loss-maker.

# Part 2 -- Why use NTMs at all?

## The political-economy answer {.m}

- Tariffs sit in a public schedule -- consumers, downstream users and trade partners can read them.

- NTMs are **opaque**:
    - Tariff equivalents must be estimated, often product-by-product.
    - Compliance cost is borne by the firm, not measured in the budget.

- Concentrated benefits, dispersed costs (Olson, 1965): the gainers organise, the losers do not.

- Indonesia food-NTMs estimated **~49% tariff-equivalent** ([The Conversation, 2022](https://theconversation.com/jangan-dibatasi-impor-bahan-makanan-bisa-perkuat-perekonomian-dan-ketahanan-pangan-192625)) -- a number no parliament would approve as an explicit tariff.

## The economic-rationale answer {.m}

Some NTMs address **real market failures**:

- **Information** -- product standards, labelling.

- **Externalities** -- emissions, pesticide residue, financial-stability spillovers. Carbon Border Adjustment (EU CBAM) is the canonical recent case.

- **Strategic competitiveness / appropriability** -- subsidies for R&D-intensive industries where private investment under-delivers.

- **Resilience** -- post-COVID supply-chain redundancy; pandemic stockpiles; semiconductors.

The honest version of the infant-industry argument lives here too. The catch from last week still applies: a domestic policy is almost always a better instrument than a trade policy for a domestic problem.

## The geopolitical answer (post-2018) {.m}

- US export controls on advanced semiconductors to China (2022 Oct, 2023 Oct).

- Inward FDI screening: CFIUS (US), EU FDI Regulation, UK NSI Act, Indonesia DNI revisions.

- "Friend-shoring", "de-risking", critical-minerals partnerships.

- US 2025 tariff package (broad, escalatory, partly outside WTO process).

- The policy paradigm has tilted from **"open by default, intervene if needed"** to **"managed by default, liberalise if convenient"**. The 2022-25 distortive count in GTA confirms it.

# Part 3 -- The World Trade Organization

## Why multilateral rules? {.m}

The cautionary tale of the 1930s:

- **Smoot-Hawley Tariff Act (US, 1930)** -- average tariff on dutiable imports raised to ~59%.

- Retaliation -- Canada, France, UK, others raised their tariffs in turn.

- World trade collapsed by roughly **two-thirds, 1929-1933**.

- Beggar-thy-neighbour devaluations and import controls deepened the Depression.

- Lesson taken at Bretton Woods (1944): **tie hands collectively** so no single government's domestic politics can unleash a global trade war.

## GATT 1947: a "provisional" treaty that lasted 47 years {.m}

- Three Bretton Woods siblings designed: IMF (money), World Bank (reconstruction), **ITO (trade)**.

- ITO Charter (Havana, 1948) was rejected by the US Senate in 1950 $\rightarrow$ ITO never born.

- Its tariff-negotiation chapter survived as the **General Agreement on Tariffs and Trade (GATT)** -- signed by 23 contracting parties, applied "provisionally" from 1 January 1948.

- "Provisionally" until 1995.

## GATT core principles {.m}

Four principles still anchor today's WTO:

1. **Most-Favoured Nation (Art. I).** Any concession granted to one member is granted to *all* members. (Exceptions: FTAs, GSP, balance-of-payments.)

2. **National Treatment (Art. III).** Once an import has cleared customs, it must be treated no worse than a like domestic product (taxes, regulations).

3. **Tariff bindings.** Each member has a *schedule* of bound (ceiling) rates; applied rates can be lower but not higher.

4. **Reciprocity.** Concessions are exchanged, not given away -- liberalisation is a negotiated political deal.

## From GATT rounds to the WTO {.m}

- 8 GATT rounds, 1947-1994. Early rounds: tariff cuts on industrial goods.

- **Tokyo Round (1973-79):** first serious push on non-tariff measures -- voluntary "codes" on subsidies, AD, standards, government procurement, customs valuation.

- **Uruguay Round (1986-94):** the big bang.
    - Agriculture brought in (AoA).
    - Textiles brought in (ATC, phase-out of MFA).
    - Services (GATS), intellectual property (TRIPS), TRIMs.
    - Strengthened, binding **dispute settlement** (DSU).
    - **WTO created, in force 1 January 1995.**

## What the WTO actually is {.m}

- **Member-driven.** 164 members, ~98% of world trade. Indonesia is a founding member.

- Three legs:
    - **GATT** (goods), **GATS** (services), **TRIPS** (intellectual property).

- Plus a **Dispute Settlement Body** -- panels + Appellate Body, with binding rulings and authorised retaliation. For 25 years, the "crown jewel".

- No real legislative capacity: change the rules only by member consensus.

## Doha "Development" Round: the agenda {.s}

- Launched **November 2001, Doha Ministerial** -- two months after 9/11, framed as a deliberate signal that the world could still cooperate.

- Called the **"Development Round"** in response to developing-country complaints about the Uruguay outcomes:
    - TRIPS pushed up the cost of medicines and technology.
    - Agriculture liberalisation in rich countries had barely begun.
    - "Implementation issues" -- promised technical assistance never materialised.

- Built on the **single undertaking** -- "nothing is agreed until everything is agreed" -- forcing cross-issue trade-offs.

- Eight negotiating tracks: agriculture, NAMA (industrial tariffs), services (GATS), TRIPS & public health, Singapore issues (investment, competition, procurement, trade facilitation), rules (AD/SCM, FTAs), environment, special & differential treatment.

## Doha: the slow collapse, 2003-2017 {.s}

- **Cancún 2003 (MC5):** collapsed. Singapore issues + the **cotton initiative** (4 West African states demanding the end of US cotton subsidies). Birth of the **G20 trade bloc** -- Brazil, India, South Africa, China -- as a counterweight to the US-EU duopoly.

- **Hong Kong 2005 (MC6):** agreed to phase out agricultural *export* subsidies.

- **Geneva July 2008 mini-ministerial:** very close to a deal. Broke down over the **Special Safeguard Mechanism** (SSM): India insisted on a low trigger to protect subsistence farmers from import surges; US refused. One issue, one deal-breaker.

- **Bali 2013 (MC9):** salvaged the **Trade Facilitation Agreement** -- first multilateral deal *born* under the WTO.

- **Nairobi 2015 (MC10):** end of agricultural export subsidies formally agreed, but the US and EU **declined to recommit to the Doha mandate**. *Post-Doha* era begins.

- Three deeper reasons for failure: (i) single undertaking made a deal infeasible; (ii) **China's 2001 accession** transformed the negotiating arithmetic; (iii) the US pivoted to mega-regionals.

## Plan B: the rise of FTAs {.m}

- Members responded by going **bilateral / regional**:
    - CPTPP (2018), USMCA (2020), RCEP (2022), AfCFTA (2021), IPEF (2022).
    - EU FTA pipeline (Mercosur, Indonesia IEU-CEPA, Vietnam, Korea, Japan).

- Plurilateral **Joint Statement Initiatives** inside the WTO: e-commerce, services domestic regulation, investment facilitation.

- For Indonesia: founder of ASEAN Economic Community; member of RCEP; IEU-CEPA after 19 negotiation rounds (2016-2025).

- Cost of multilateral failure: a **"spaghetti bowl"** of overlapping FTAs with different rules of origin -- compliance burden falls on firms.

## The WTO's current crisis {.m}

- **Appellate Body paralysed since 11 December 2019.** US (under both parties) has blocked new appointments since 2017, citing judicial overreach.

- Without a functioning AB, a losing party can **appeal "into the void"** and the ruling never becomes binding.

- Workaround: 26 members (incl. EU, China, Australia) joined the **Multi-Party Interim Appeal Arrangement (MPIA)**. Indonesia and the US have *not* joined.

- US 2025 tariff package proceeded largely outside WTO-consistent process.

- The discipline that took 50 years to build is being eroded faster than it can be defended.

## Indonesia at the WTO {.m}

- **DS592 -- Indonesia (Nickel ore export ban), 2019 -.**
    - EU complaint against the 2020 ore-export ban and related domestic-processing requirements.
    - Panel report (Nov 2022): Indonesia's measures inconsistent with GATT Art. XI:1 (quantitative restrictions); justifications under Art. XX (general exceptions) rejected.
    - Indonesia appealed -- *into the void.* No binding outcome. The ban remains in force.

- Other live or recent disputes: chicken (DS484, won by Brazil), bioethanol AD, several palm-oil disputes against the EU.

- Indonesia is both a beneficiary of the rules and a frequent test case for their limits.

# Part 4 -- What governments actually use

## Global Trade Alert (GTA / NIPO) {.m}

- Independent monitoring database of unilateral trade and industrial policy interventions, run from St. Gallen.

- **Scope:**
    - Coverage starts 1 November 2008.
    - **Discriminatory** measures only -- a uniform VAT change is excluded.
    - **Unilateral** -- FTAs, RCEP, EU accession schedules are *not* in here.
    - Not WTO-notified TBT/SPS, not UN sanctions, not CITES.
    - Subsidies below USD 10M (USD 100M for SME-only) excluded.

- Each record is coded by instrument, sector, affected partner, and an "initial assessment" (distortive vs liberalising).

- Used today: April 2026 inventory, **53,102 measures**.

## Most-used instruments, 2009--2025

```{r top-tools}
#| fig-width: 10
#| fig-height: 5.2
top_tools <- gta |>
  filter(!is.na(type), year >= 2009, year <= 2025) |>
  count(type, sort = TRUE) |>
  slice_head(n = 12) |>
  mutate(type = str_wrap(type, 30),
         type = forcats::fct_reorder(type, n))

ggplot(top_tools, aes(n, type)) +
  geom_col(fill = "#26351c") +
  geom_text(aes(label = comma(n)), hjust = -0.1, size = 4) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0, 0.12))) +
  labs(x = "Number of GTA measures", y = NULL,
       title = "Top 12 trade-policy instruments, world, 2009-2025",
       caption = "Source: Global Trade Alert / NIPO, April 2026 inventory.") +
  den_theme
```

::: {.s}
Tariffs lead, but **subsidies** (financial grant + state loan + state aid + trade finance) dominate when summed. Anti-dumping is the largest single trade-defence tool.
:::

## Distortive vs liberalising over time

```{r world-trend}
#| fig-width: 10
#| fig-height: 5.2
trend <- gta |>
  filter(year >= 2009, year <= 2025,
         assess %in% c("Distortive", "Liberalising")) |>
  count(year, assess)

ggplot(trend, aes(year, n, fill = assess)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Distortive" = "#a23b2c", "Liberalising" = "#26351c"),
                    name = NULL) +
  scale_x_continuous(breaks = seq(2009, 2025, 2)) +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Number of measures",
       title = "World trade-policy interventions, by initial assessment",
       caption = "Source: Global Trade Alert / NIPO, April 2026.") +
  den_theme +
  theme(legend.position = "top")
```

::: {.s}
Inflection points: 2009 (post-GFC), 2018 (US-China), 2020 (COVID), 2022+ (industrial-policy revival -- IRA, CHIPS, EU CBAM).
:::

## Who intervenes most?

```{r top-impl}
#| fig-width: 10
#| fig-height: 5.2
top_impl <- gta |>
  filter(year >= 2009, year <= 2025) |>
  count(impl, sort = TRUE) |>
  slice_head(n = 15) |>
  mutate(impl = forcats::fct_reorder(impl, n))

ggplot(top_impl, aes(n, impl)) +
  geom_col(fill = "#51483d") +
  geom_text(aes(label = comma(n)), hjust = -0.1, size = 4) +
  scale_x_continuous(labels = comma, expand = expansion(mult = c(0, 0.12))) +
  labs(x = "Number of measures (any assessment)", y = NULL,
       title = "Top 15 implementing jurisdictions, 2009-2025",
       caption = "Source: Global Trade Alert / NIPO, April 2026.") +
  den_theme
```

::: {.s}
**The big interveners are big and rich, not poor.** Industrial policy is a rich-country game; developing countries follow.
:::

## Indonesia: top instruments

```{r idn-tools}
#| fig-width: 10
#| fig-height: 5.2
idn_tools <- gta |>
  filter(impl == "Indonesia", !is.na(type)) |>
  count(type, sort = TRUE) |>
  slice_head(n = 12) |>
  mutate(type = str_wrap(type, 32),
         type = forcats::fct_reorder(type, n))

ggplot(idn_tools, aes(n, type)) +
  geom_col(fill = "#26351c") +
  geom_text(aes(label = n), hjust = -0.2, size = 4) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = "Number of measures", y = NULL,
       title = "Indonesia: top 12 instruments since 2009",
       caption = "Source: Global Trade Alert / NIPO, April 2026.") +
  den_theme
```

::: {.s}
Note the **export side** -- export tax, export price benchmark, export ban, export licensing. Distinctive for a resource-exporting economy.
:::

## Indonesia: pace of intervention

```{r idn-trend}
#| fig-width: 10
#| fig-height: 5.0
idn_trend <- gta |>
  filter(impl == "Indonesia", year >= 2009, year <= 2025,
         assess %in% c("Distortive", "Liberalising")) |>
  count(year, assess)

ggplot(idn_trend, aes(year, n, fill = assess)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("Distortive" = "#a23b2c", "Liberalising" = "#26351c"),
                    name = NULL) +
  scale_x_continuous(breaks = seq(2009, 2025, 2)) +
  labs(x = NULL, y = "Measures implemented",
       title = "Indonesia: implemented measures by year",
       caption = "Source: Global Trade Alert / NIPO, April 2026.") +
  den_theme +
  theme(legend.position = "top")
```

::: {.s}
Of 502 Indonesian measures since 2009: ~66% distortive, ~34% liberalising. Notable: 2009 commodity-boom export controls; 2014 mineral-ore export ban; 2020 nickel ore ban; 2024-25 hilirisasi continues.
:::

# Wrap-up

## Take-aways {.m}

1. **Tariffs are visible. The real action is everywhere else** -- subsidies, LCRs, standards, contingent protection, export controls.

2. NTMs have **legitimate uses** (information, externalities, resilience) and **high abuse potential** (opacity, rent-seeking).

3. The **WTO** was the post-war answer to 1930s-style trade collapse. Its *negotiation* function stalled in 2008; its *adjudication* function broke in 2019; its *rule-book* still binds, weakly.

4. For Indonesia the question is not *whether* to intervene -- we already do, ~30 measures a year on average -- but **whether to intervene well or badly**. That is what this course is for.

## Next week

**Balance of Payments and the data of international economics.**

- The current account, the financial account, why the BOP must balance.

- Reading data: BPS, Bank Indonesia SEKI, IMF BOP, World Bank.

- Hand-on: pulling and plotting Indonesia's BOP series.
