---
title: "Ricardian Model"
subtitle: "ECES905205 pertemuan 2"
date: "5 September 2022"
author: "I Made Krisna Gupta"
format: 
  revealjs:
    theme: [simple,../mytheme.scss]
    chalkboard: true
    incremental: false
    transition: convex
    background-transition: fade
    slide-number: true
  beamer: default

editor: visual
---

## Today

-   Comparative advantage
-   Ricardian model sebagai pondasi *trade theory*

## Konsep Comparative Advantage {.m}

-   Indonesia adalah negara agraris. Kenapa impor kedelai?

    -   "... (profit menanam kedelai) jauh di bawah keuntungan yang didapat dari menanam padi dan jagung." ([kompas](https://money.kompas.com/read/2021/01/14/050900426/menurut-kementan-ini-penyebab-petani-enggan-menanam-kedelai-?page=all))

-   Sementara itu, "Indonesian consumers have developed a preference for U.S. soybeans due in part to their uniform size, color and suitability for tempeh and tofu manufacturing." ([ussoy.org](https://ussoy.org/growth-in-tempeh-market-offers-global-opportunities/))

-   tempe menjadi produk ekspor andalan di [Jepang](https://en.tempo.co/read/1512185/13-tons-tempeh-chips-exported-to-japan) dan [Amerika Serikat](https://ussoy.org/growth-in-tempeh-market-offers-global-opportunities/)

-   what does it tell you about the structure of the Indonesian economy?

-   what does it tell you about *comparative advantage*?

## Ekonomi dengan 1 faktor produksi

-   1 faktor produksi: buruh.

-   2 barang: Apel dan Beras.

-   Teknologi direpresentasikan dengan *unit labor requirement*

    -   berapa jumlah jam kerja yang diperlukan untuk menghasilkan 1 unit barang.
    -   cara ini kebalikan dari tipikal produktivitas: angka besar berarti inefisien.
    -   $a_{A}$ untuk apel, $a_{B}$ untuk beras.

## Ekonomi dengan 1 faktor produksi

-   Total *labour supply* $= L$

-   $a_{A}Q_A =$ total jam kerja yang diperlukan untuk menghasilkan apel sebanyak $Q_A$

-   $a_{B}Q_B =$ total jam kerja yang diperlukan untuk menghasilkan beras sebanyak $Q_B$

-   Production constraint:

$$
a_{A}Q_A + a_{B}Q_B \leq L
$$

## PPF

::: {.s layout-ncol="2"}
-   Bayangkan sebuah negara (sebut saja H):
    -   Butuh 1 OJ untuk membuat apel, 2 OJ untuk membuat beras.
    -   Total OJ di negara H adalah 1.000 OJ $(L=1000)$
    -   Jika semua OJ digunakan untuk apel, maka $A,B=\{1000,0\}$
    -   Jika semua OJ digunakan untuk beras, maka $A,B=\{0,500\}$
-   Linear PPF $\Rightarrow$ opportunity cost konstan.
    -   Memutuskan bikin beras 1kg = memutuskan gak bikin apel 2kg.
    -   Memutuskan bikin apel 1kg = memutuskan gak bikin beras $\frac{1}{2}$kg.

```{r}
#| fig-width: 5
#| fig-height: 4
#| label: fig1
#| fig-cap: "Production Possibility Frontier (PPF) negara H"
library('ggplot2')
graph<-ggplot()+
  theme_classic()

home <- function(x) 500-0.5*x
graph+
  stat_function(fun = home, size=1.1)+
  scale_y_continuous(expand=c(0,0),limits=c(0,1000), breaks = c(0,500,1000))+
  scale_x_continuous(expand=c(0,0),limits=c(0,1000), breaks =c(0,500,1000))+
  labs(x="Apel (kg)",y="Beras (kg)")
```
:::

$$
1Q_A + 2Q_B \leq 1000
$$

## PPF and Relative Prices

::: {.columns .s}
::: {.column width="48%"}
A simple labor theory of value:

> In the absence of international trade, the relative prices of goods are equal to their relative unit labor requirements.

-   if $P_A=P_B$, better concentrate on making A:
    -   wage $={1000\times P_A} \geq 500 \times P_B$
-   workers are better off making B only if $P_A < \frac{1}{2}P_B$
    -   workers only produce both iff $P_A = \frac{1}{2}P_B$
-   equilibrium is at $\frac{a_A}{a_B}=\frac{P_A}{P_B}$
:::

::: {.column width="48%"}
```{r}
#| fig-width: 5
#| fig-height: 4
#| label: fig3
#| fig-cap: "Production Possibility Frontier (PPF) negara H"
library('ggplot2')
graph<-ggplot()+
  theme_classic()

home <- function(x) 500-0.5*x
graph+
  stat_function(fun = home, size=1.1)+
  scale_y_continuous(expand=c(0,0),limits=c(0,1000), breaks = c(0,500,1000))+
  scale_x_continuous(expand=c(0,0),limits=c(0,1000), breaks =c(0,500,1000))+
  labs(x="Apel (kg)",y="Beras (kg)")
```
:::
:::

## PPF and Trade

-   To understand this term better, we will use some simplified hypothetical examples.

-   Suppose there are two economies, IDN and USA, and two goods, textiles (T) and soybean (S).

```{=tex}
\begin{equation} \label{eq1}
\begin{split}
IDN: \ 1T+4S&=200 \\
USA: \ 2T+2S & = 200 
\end{split}
\end{equation}
```
::: s
## 2 countries in autarky

```{=tex}
\begin{equation} \label{eq2}
\begin{split}
IDN: \ 1T+4S&=200 \\
USA: \ 2T+2S & = 200 
\end{split}
\end{equation}
```
| Produce           | US opportunity cost | IDN opportunity cost |
|-------------------|---------------------|----------------------|
| 1 ton of textiles | 1 ton of soybean    | 250 kg of soybean    |
| 1 ton of soybean  | 1 ton of textiles   | 4 ton of soybean     |

-   IDN is comparatively better at making textiles, while US is better at producing soybean:
    -   To make 1 ton of textiles, US needs to gave up resources which can be used to make 1 ton soybean, while Indonesians need only to give up 250 kg soybean.
    -   To make 1 ton of soybean, US needs to gave up resources which can be used to make 1 ton textiles, while Indonesians need only to give up 4 ton of textiles.
-   Note:

$$
\frac{a_{it}}{a_{is}}=\frac{1}{4} \geq \frac{1}{1}=\frac{a_{ut}}{a_{us}}
$$
:::

------------------------------------------------------------------------

Suppose the two countries do not trade. Indonesia's production is {T,S}={100,25}, While in the US, it's {50,50}.

```{r grafik1,echo=FALSE, message=FALSE, warning=FALSE}
#| fig-width: 7
#| fig-height: 5.5
#| label: fig2

library(patchwork)
library(hrbrthemes)
library(kableExtra)
USA <- function(x) 100-x
IDN <- function(x) 50-.25*x

amrik <- graph +
  stat_function(fun = USA, color="blue", size=1.1)+
  scale_y_continuous(expand=c(0,0),limits=c(0,100), breaks = c(50,100))+
  scale_x_continuous(expand=c(0,0),limits=c(0,200), breaks =c(50,100))+
  labs(x="Quantity of textiles (ton)",y="Quantity of soybeans (ton)",
       title = "USA PPF",size=1)+
  annotate("label", x = 85, y = 70, label = 'US production \n and consumption \n in autarky' ,size=5)+
  geom_segment(aes(x = 50, y = 50, xend = 70, yend = 61)
               ,size=1)+
  geom_point(aes(x=50,y=50),color='black',size=4)+
  geom_segment(aes(x = 0, y = 50, xend = 50, yend = 50), lty = "dotted"
               ,size=1) +
  geom_segment(aes(x = 50, y = 0, xend = 50, yend = 50), lty = "dotted"
               ,size=1)
  
indon <- graph +
  stat_function(fun = IDN, color="red", size=1.1)+
  scale_y_continuous(expand=c(0,0),limits=c(0,100), breaks = c(50,25))+
  scale_x_continuous(expand=c(0,0),limits=c(0,200), breaks =c(100,200))+
  labs(x="Quantity of textiles (ton)",y="Quantity of soybeans (ton)",
       title = "IDN PPF",size=1)+
  annotate("label", x = 85, y = 60, label = 'IDN production \n and consumption \n in autarky' ,size=5)+
  geom_segment(aes(x = 100, y = 25, xend = 70, yend = 51)
               ,size=1)+
  geom_point(aes(x=100,y=25),color='black',size=4)+
  geom_segment(aes(x = 0, y = 25, xend = 100, yend = 25), lty = "dotted"
               ,size=2) +
  geom_segment(aes(x = 100, y = 0, xend = 100, yend = 25), lty = "dotted"
               ,size=2)

indon+amrik
```

## Gains from trade

```{r tab,echo=FALSE, message=FALSE, warning=FALSE}
game <- data.frame(
  "country"=c(rep('USA',times=2),rep('IDN',times=2)), "goods"=c("textiles (ton)","soybean (ton)","textiles (ton)","soybean (ton)"),
  "p1"=c(50,50,100,25), "p2"=c(50,50,100,25),"p3"=c(0,100,200,0),
  "p4"=c(75,62.5,125,37.5),"gains"=c("+25","+12.5","+25","+12.5")
)

zzz<-kbl(game, align = "c", format='html',
    col.names = c(" "," ", "production","consumption",
                     "production","consumption", "Gains from trade")) %>%
  kable_styling(font_size=18) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1) %>%
  add_header_above(c(" "," ", "in autarky" = 2,"with trade"=3))
zzz
```

By concentrating production on one good and then trade, you can see that both ended up better-off:

-   with autarky, total production {textiles,soybean} $=\{150,75\}$

-   consumption=production, and $C_{USA} = \{50,50\},C_{IDN} = \{100,25\}$

------------------------------------------------------------------------

## Gains from trade

```{r tab1,echo=FALSE, message=FALSE, warning=FALSE}
zzz
```

-   with trade, total production {textiles,soybean} $=\{200,100\}$

    -   Production $\Rightarrow Q_{USA}=\{0,100\}, Q_{IDN}=\{200,0\}$
    -   Consumption $\Rightarrow C_{USA}=\{75,62.5\}, C_{IDN}=\{125,37.5\}$

------------------------------------------------------------------------

```{r grafik2,echo=FALSE, message=FALSE, warning=FALSE,fig.width=10,fig.height=8}

amrik1 <- graph +
  stat_function(fun = USA, color="blue", size=1.1)+
  scale_y_continuous(expand=c(0,0),limits=c(0,100), 
                     breaks = c(50,62.5,100))+
  scale_x_continuous(expand=c(0,0),limits=c(0,200), 
                     breaks =c(50,75,100))+
  labs(x="Quantity of textiles (ton)",y="Quantity of soybeans (ton)",
       title = "USA PPF",size=0.5)+
  annotate("label", x = 130, y = 30, label = 'US production \n and consumption \n in autarky' ,size=5)+
  annotate("label", x = 65, y = 90, label = 'US production \n with trade' ,size=5)+
  annotate("label", x = 130, y = 62.5, label = 'US consumption \n with trade' ,size=5)+
  geom_segment(aes(x = 0, y = 100, xend = 35, yend = 90 )
               ,size=1)+
  geom_segment(aes(x = 75, y = 62.50, xend = 95, yend = 63)
               ,size=1)+
  geom_segment(aes(x = 50, y = 50, xend = 100, yend = 40)
               ,size=1)+
  geom_point(aes(x=50,y=50),color='black',size=4)+
  geom_point(aes(x=75,y=62.5),color='black',size=4)+
  geom_point(aes(x=0,y=100),color='black',size=4)+
  geom_segment(aes(x = 0, y = 50, xend = 50, yend = 50), lty = "dotted"
               ,size=1) +
  geom_segment(aes(x = 50, y = 0, xend = 50, yend = 50), lty = "dotted"
               ,size=1)+
  geom_segment(aes(x = 0, y = 62.5, xend = 75, yend = 62.5), lty = "dotted"
               ,size=1) +
  geom_segment(aes(x = 75, y = 0, xend = 75, yend = 62.5), lty = "dotted"
               ,size=1)

indon1 <- graph +
  stat_function(fun = IDN, color="red", size=1.1)+
  scale_y_continuous(expand=c(0,0),limits=c(0,100), breaks = c(50,37.5,25))+
  scale_x_continuous(expand=c(0,0),limits=c(0,200), breaks =c(100,125,200))+
  labs(x="Quantity of textiles (ton)",y="Quantity of soybeans (ton)",
       title = "IDN PPF",size=1)+
  annotate("label", x = 45, y = 80, 
           label = 'IDN production \n and consumption \n in autarky' ,
           size=5)+
  annotate("label", x = 95, y = 65,
           label = 'IDN consumption \n with trade' ,size=5)+
  annotate("label", x = 160, y = 30, 
           label = 'IDN production \n with trade' ,size=5)+
  geom_segment(aes(x = 100, y = 25, xend = 35, yend = 73)
               ,size=1)+
  geom_segment(aes(x = 125, y = 37.5, xend = 90, yend = 59)
               ,size=1)+
  geom_segment(aes(x = 200, y = 0, xend = 160, yend = 25)
               ,size=1)+
  geom_point(aes(x=100,y=25),color='black',size=4)+
  geom_point(aes(x=125,y=37.5),color='black',size=4)+
  geom_point(aes(x=200,y=0),color='black',size=4)+
  geom_segment(aes(x = 0, y = 25, xend = 100, yend = 25), lty = "dotted"
               ,size=1) +
  geom_segment(aes(x = 100, y = 0, xend = 100, yend = 25), lty = "dotted"
               ,size=1)+
  geom_segment(aes(x = 0, y = 37.5, xend = 125, yend = 37.5), lty = "dotted"
               ,size=1) +
  geom_segment(aes(x = 125, y = 0, xend = 125, yend = 37.5), lty = "dotted"
               ,size=1)
indon1+amrik1
```

------------------------------------------------------------------------

```{r}
#| fig-width: 5
#| fig-height: 4
#| label: fig4

graph<-ggplot()+
  theme_classic()
USA <- function(x) 100-x
IDN <- function(x) 50-.25*x
gab <- function(x) 100-.5*x
graph+
  stat_function(fun = gab, size=1.1)+
  stat_function(fun = USA, size=1.1,color="red")+
  stat_function(fun = IDN, size=1.1,color="blue")+
  scale_y_continuous(expand=c(0,0),limits=c(0,100), breaks = c(0,50,100))+
  scale_x_continuous(expand=c(0,0),limits=c(0,200), breaks =c(0,100,200))+
  labs(x="Quantity of Textiles (ton)",y="Quantity of Soybean (ton)")+
  geom_point(aes(x=100,y=25),color='black',size=4)+
  geom_point(aes(x=125,y=37.5),color='black',size=4)+
  geom_point(aes(x=75,y=62.5),color='black',size=4)+
  geom_point(aes(x=50,y=50),color='black',size=4)+
  annotate("label", x = 100, y = 30, label = 'CA[IDN]' ,size=5,parse=TRUE)+
  annotate("label", x = 125, y = 45, label = 'CT[IDN]' ,size=5,parse=TRUE)+
  annotate("label", x = 75, y = 70, label = 'CT[USA]' ,size=5,parse=TRUE)+
  annotate("label", x = 50, y = 55, label = 'CA[USA]' ,size=5,parse=TRUE)

```

------------------------------------------------------------------------

## Comparative advantage

-   In reality, we will need country's preferences to establish proper equilibrium quantity and prices.

-   In this model, we have no price. But from the opportunity cost, we can calculate the relative prices:

    -   In the US, opportunity cost to produce 1 soybean is 1 textile. That means, $P^{soybean}_{USA} = P^{textile}_{USA}$
    -   In IDN, opportunity cost to make 1 soybean is 4 textiles. That means, the price of soybean must be four times price of textile.

## Comparative advantage

-   Note that in autarky, both countries consume (and produce) with the same proportion as their unit labor cost.

-   With trade, both countries produce with 1:2 proportion

    -   their consumptions are skewed toward 1:2

-   This reflects the change in relative prices:

    -   with trade, textile gets relatively cheaper in US.
    -   in IDN, soybean gets a bit cheaper (from 1:4 to around 1:3.33)

------------------------------------------------------------------------

## Comparative VS Absolute advantage {.m}

-   It's possible that a country is good at producing many things.
    -   This is called absolute advantage.
-   There's no reason why US can't be better at making clothes than Indonesia.
    -   Hence, it's possible that in fact, US still better at making clothes
-   But what matters is *relative* opportunity cost:
    -   as long as its better for US capital and labour to produce soybean, they will still buy clothes from Indonesia.
    -   It may take lots of labor to produce textiles in Indonesia, but it'll take even more to produce soybean.
-   Makin specialised, makin besar opportunity cost.
    -   S2 UI vs SMEA.

------------------------------------------------------------------------

## Absolute advantage

-   Suppose USA dominates IDN in both:

```{=tex}
\begin{equation} \label{eq3}
\begin{split}
IDN: \ 2.5T+4S&=200 \\
USA: \ 2T+2S & = 200 
\end{split}
\end{equation}
```
-   Autarky: $Q_{USA}=\{50,50\}, Q_{IDN}=\{40,25\}$

-   Trade: $Q_{USA}=\{10,90\}, Q_{IDN}=\{80,0\}$

-   Total: $Q_{autarky}=\{90,75\}, Q_{trade}=\{90,90\}$

------------------------------------------------------------------------

## A note on absolute advantage

-   Mungkin anda menyadari bahwa absolute advantage hanyalah versi alternatif dari Walras Law.

-   Dengan kata lain, meski ada keunggulan absolut, *contract curve* & *better allocation* tetap ada (hence pareto improvement).

    -   Meski demikian, negara yang punya keunggulan absolut tetap memiliki posisi tawar yang lebih baik.

-   *trade is pareto improving* in the sense that we compare country A before trade with country A after trade.

    -   we do not compare country A with country B.

------------------------------------------------------------------------

## Taking from the poor?

-   In short, because IDNsians are less productive compared to the US, it's better to US to buy cheaper goods to IDN.

-   This is not to say that US take advantage from IDN:

    -   Since IDN labour is less productive anyway, they will still have low wage even if they sell domestically.
    -   In fact, serving US market may give them higher wage.

-   For example, if you feel ordering go-food is taking advantage from tukang ojeg, the alternative is even worse:

    -   if you ended up stop ordering go-food, their income is gone.

------------------------------------------------------------------------

## What determines comparative advantage? {.m}

1.  **Differences in climate** is the reason why we're so good at producing CPO and rubber, but sucks at producing soybean and wheat.

2.  **Differences in Factor Endowment**. Some countries are endowed with natural resource, some with cheap labour. Countries which has no both has to find something else, such as:

3.  **Differences in technology**. Japan, South Korea and Taiwan are one good example. While technology can be transferred, opportunity cost of investing in high-tech things is more production of CPOs.

------------------------------------------------------------------------

## Trade is Good {.m}

-   Barusan kita mempelajari bahwa ada gains dari trade bahkan ketika trading partner punya absolute advantage atas semua produksi.

-   Jika tidak ada pun, *autarky* akan terjadi tanpa dipasang penghalang / *barier*

    -   ini gak papa jika *autarky* terjadi tanpa pemaksaan.

-   Untuk analisis yang menyeluruh, kita perlu memperhitungkan *preference*

    -   *preference* dapat direfleksikan dengan harga.
    -   namun intuisinya tidak berbeda jauh.
    -   kita bahas di pertemuan-pertemuan berikutnya.

## Multi-goods comparative advantage

-   Misalnya ada N goods.

-   unit labor cost di home jadi $a_i$ di mana $i \in (1,N), N>2$

    -   correspondingly, $a_i^*$ adalah untuk foreign

-   untuk setiap good $i$, kita buat $\frac{a_i}{a_i^*}$

-   lalu urutkan sehingga $\frac{a_1}{a_1^*} < \frac{a_2}{a_2^*} < ... < \frac{a_N}{a_N^*}$

    -   alias diurutkan berdasarkan comparative advantage.

## Multi-goods comparative advantage

-   Let $w$ be home wage per OJ, while $w^*$ be foreign wage.
    -   artinya, biaya membuat 1 apel di home $= wa_{apel}$
    -   biaya membuat 1 apel di foreign $= w^*a_{apel}^*$
-   Goods are produced in the cheapest place.
    -   Apel dibikin di home jika: $wa_{apel}<w^*a_{apel}^*$
    -   atau jika: $\frac{a_{apel}^*}{a_{apel}}>\frac{w}{w^*}$

## Multi-goods comparative advantage {.m}

-   Let 2 economies with this characteristics:

| Good $(i)$ | $a_i$ | $a_i^*$ | $\frac{a_i^*}{a_i}$ |
|------------|:-----:|:-------:|:-------------------:|
| Apel       |   1   |   10    |         10          |
| Beras      |   5   |   40    |          8          |
| CPO        |   3   |   12    |          4          |
| Durian     |   6   |   12    |          2          |
| Emas       |  12   |    9    |        0.75         |

-   Note that table above is sorted decendingly by $\frac{a^*_i}{a_i}$

-   or: $\frac{a_1}{a_1^*} < \frac{a_2}{a_2^*} < ... < \frac{a_5}{a_5^*}$

## Multi-goods comparative advantage {.m}

| Good $(i)$ | $a_i$ | $a_i^*$ | $\frac{a_i^*}{a_i}$ |
|------------|:-----:|:-------:|:-------------------:|
| Apel       |   1   |   10    |         10          |
| Beras      |   5   |   40    |          8          |
| CPO        |   3   |   12    |          4          |
| Durian     |   6   |   12    |          2          |
| Emas       |  12   |    9    |        0.75         |

-   Ingat: dibikin di home jika: $\frac{a_{i}^*}{a_{i}}>\frac{w}{w^*}$

-   say $\frac{w}{w^*}=\frac{12}{4}=3$

-   Maka apel, beras dan CPO akan dibuat domestik.

-   Durian dan Emas diimpor.

## How useful?

-   Relative productivity dan relative wage merupakan poin analisis yang sangat intuitif.
    -   gets complicated very soon as we add countries and factors.
    -   tastes, economic structures, etc
    -   but the intuition is very similar.
-   Namun jika analisis ini benar, berarti negara-negara akan auto-spesialisasi? Apakah ini yang kita observe di dunia nyata?

## What prevents specialisation?

-   More than 1 factors
-   Protectionisms.
    -   more on this later.
-   Transportation cost.
    -   Some places are inherently difficult to reach.
    -   Some goods are harder to move.
-   at the extreme, a good can be **nontradable good**.
    -   most services are like this.

------------------------------------------------------------------------

## What prevents specialisation?

-   Ricardian model: 1 very mobile factor of production.

-   In reality, trade matters in distribution of income.

    -   short-run: moving factors are costly.
    -   long-run: shift in factor demand.

-   Factor owners (land, labor, capital, etc) receive different gains from trade!

## Japanese rice farmer case:

-   Land scarcity -\> rice is expensive (even compared to US).
-   Free trade means higher real wage.
-   Farmers may need to move industries.
    -   reskilling is costly.
    -   land value drops.
    -   machines & infrastructure become useless.
-   Better spend money to organize politically!

## Specific factors

-   Next, we add one more factor to the model: a specific factor.

-   Labor can move between sectors, but this one other factor is specific to the sector, hence can't move freely.

    -   just like our Japanese farmers!

-   We will also discuss a bit what is this specific factor of production in reality.

------------------------------------------------------------------------

::: r-stack
<br/>

<br/>

<iframe src="https://giphy.com/embed/3ohs86vZAWiJXWvQI0" width="480" height="353" frameBorder="0" class="giphy-embed" allowFullScreen>

</iframe>

<p><a href="https://giphy.com/gifs/eminem-my-name-is-the-slim-shady-lp-3ohs86vZAWiJXWvQI0">via GIPHY</a></p>
:::
